---
title: "SpE data exploration"
##output: html_notebook
---

```{r}
library("SpatialExperiment")
library("here")
library(qs2)
here::i_am('.git/HEAD')
datadir=here("processed-data", "rds")
refdir=here("processed-data", "ref")
```

```{r}
x <- readRDS(file.path(datadir, "PB_dx_spg", "pseudo_neun_pos_donor_spd.rds"))
## Neuronal SPG (microenvironment) context, pseudobulked by Spatial Domain (SpD)
dim(x)
names(colData(x))
#head(colData(x)[, c("age", "sex", "brnum", "slide_id")])
##
```

> dim(x)
[1] 16768   308
> names(colData(x))
 [1] "sample_id"               "in_tissue"               "array_row"               "array_col"               "X10x_graphclust"         "X10x_kmeans_10_clusters"
 [7] "X10x_kmeans_2_clusters"  "X10x_kmeans_3_clusters"  "X10x_kmeans_4_clusters"  "X10x_kmeans_5_clusters"  "X10x_kmeans_6_clusters"  "X10x_kmeans_7_clusters"
[13] "X10x_kmeans_8_clusters"  "X10x_kmeans_9_clusters"  "key"                     "sum_umi"                 "sum_gene"                "expr_chrM"
[19] "expr_chrM_ratio"         "spg_NDAPI"               "spg_PDAPI"               "spg_IDAPI"               "spg_CNDAPI"              "spg_NNeuN"
[25] "spg_PNeuN"               "spg_INeuN"               "spg_CNNeuN"              "spg_NWFA"                "spg_PWFA"                "spg_IWFA"
[31] "spg_CNWFA"               "spg_NClaudin5"           "spg_PClaudin5"           "spg_IClaudin5"           "spg_CNClaudin5"          "BrNumbr"
[37] "dx"                      "DX"                      "sex"                     "age"                     "rin"                     "slide_id"
[43] "lot_num"                 "brnum"                   "sample_label"            "outlier_cat"             "removed_spots"           "PRECAST_02"
[49] "PRECAST_03"              "PRECAST_04"              "PRECAST_05"              "PRECAST_06"              "PRECAST_07"              "PRECAST_08"
[55] "PRECAST_09"              "PRECAST_10"              "PRECAST_11"              "PRECAST_12"              "PRECAST_13"              "PRECAST_14"
[61] "PRECAST_15"              "PRECAST_16"              "fnl_spd"                 "spd_label"               "pnn_pos"                 "neuropil_pos"
[67] "neun_pos"                "vasc_pos"                "sizeFactor"              "registration_variable"   "registration_sample_id"  "ncells"



```{r}
## breakdown of Domains present for each Donor
table(x$brnum, x$spd_label)
#         SpD07-L1 SpD06-L2/3 SpD02-L3/4 SpD05-L5 SpD03-L6 SpD01-WMtz SpD04-WM
#  Br1092        1          1          1        1        1          1        0
#  Br1113        0          1          1        1        1          1        0
#  Br1139        0          0          1        1        1          0        0
#  Br1204        1          1          1        1        1          1        0
#  Br1214        0          1          1        1        1          1        1
#  Br1526        0          1          1        1        1          0        0
#  Br1578        0          1          1        1        1          1        0
#  ...
## simple count of how many domains each donor has:
table(x$brnum)
# Br1092 Br1113 Br1139 Br1204 Br1214 Br1526 Br1578 Br1682 Br1753 Br1923 Br1958 Br2039 Br2242 Br2378 Br2421 Br2470 Br2627 Br2719 Br2720 Br5182 Br5224 Br5228 Br5276 Br5314
#     6      5      3      6      6      4      5      5      5      5      5      5      5      5      5      4      6      5      4      5      6      5      5      5
# Br5367 Br5373 Br5395 Br5400 Br5436 Br5472 Br5573 Br5588 Br5590 Br5599 Br5622 Br5636 Br5639 Br5746 Br5789 Br5888 Br5928 Br5931 Br5973 Br6032 Br6257 Br6297 Br6369 Br6389
#     4      4      5      5      4      5      5      5      4      5      5      4      6      5      3      5      4      5      5      5      5      5      6      4
# Br6432 Br6437 Br6496 Br6560 Br8042 Br8207 Br8218 Br8325 Br8406 Br8433 Br8492 Br8514 Br8537 Br8667 Br8772
#      4      5      5      5      5      5      6      4      6      6      5      4      5      5      5

## which domains are generally present in this object:
sort(table(x$spd_label), decreasing=TRUE)

# SpD02-L3/4   SpD05-L5   SpD03-L6 SpD06-L2/3 SpD01-WMtz   SpD07-L1   SpD04-WM
#        63         63         62         61         47         10          2
## transition zone (47) likely has spill-over neurons from Layer 6 ?
## only 5 viable layers
```


```{r all data aggregated by domain}
d <- readRDS(file.path(datadir, "07_dx_pseudobulk", "sce_pseudo_PRECAST07_donor_spd.rds"))
dim(d) # [1] 13977   431
colnames(colData(d))

table(d$registration_variable)
# spd01 spd02 spd03 spd04 spd05 spd06 spd07
#    63    63    63    53    63    63    63

table(d$spd_label)
#  SpD07-L1 SpD06-L2/3 SpD02-L3/4   SpD05-L5   SpD03-L6 SpD01-WMtz   SpD04-WM
#        63         63         63         63         63         63         53

## white matter data only has 53 samples, but that was not unexpected given the focus on cortical layers
```

> colnames(colData(d))
 [1] "sample_id"               "in_tissue"               "array_row"               "array_col"               "X10x_graphclust"         "X10x_kmeans_10_clusters" "X10x_kmeans_2_clusters"
 [8] "X10x_kmeans_3_clusters"  "X10x_kmeans_4_clusters"  "X10x_kmeans_5_clusters"  "X10x_kmeans_6_clusters"  "X10x_kmeans_7_clusters"  "X10x_kmeans_8_clusters"  "X10x_kmeans_9_clusters"
[15] "key"                     "sum_umi"                 "sum_gene"                "expr_chrM"               "expr_chrM_ratio"         "spg_NDAPI"               "spg_PDAPI"
[22] "spg_IDAPI"               "spg_CNDAPI"              "spg_NNeuN"               "spg_PNeuN"               "spg_INeuN"               "spg_CNNeuN"              "spg_NWFA"
[29] "spg_PWFA"                "spg_IWFA"                "spg_CNWFA"               "spg_NClaudin5"           "spg_PClaudin5"           "spg_IClaudin5"           "spg_CNClaudin5"
[36] "BrNumbr"                 "dx"                      "DX"                      "sex"                     "age"                     "rin"                     "slide_id"
[43] "lot_num"                 "brnum"                   "sample_label"            "outlier_cat"             "removed_spots"           "PRECAST_02"              "PRECAST_03"
[50] "PRECAST_04"              "PRECAST_05"              "PRECAST_06"              "PRECAST_07"              "PRECAST_08"              "PRECAST_09"              "PRECAST_10"
[57] "PRECAST_11"              "PRECAST_12"              "PRECAST_13"              "PRECAST_14"              "PRECAST_15"              "PRECAST_16"              "fnl_spd"
[64] "spd_label"               "pnn_pos"                 "neuropil_pos"            "neun_pos"                "vasc_pos"                "sizeFactor"              "registration_variable"
[71] "registration_sample_id"  "ncells"


```{r gene data}
head(rowRanges(d))
```

> head(rowRanges(d))
GRanges object with 6 ranges and 7 metadata columns:
                  seqnames        ranges strand |   source     type         gene_id gene_version   gene_name      gene_type
                     <Rle>     <IRanges>  <Rle> | <factor> <factor>     <character>  <character> <character>    <character>
  ENSG00000241860     chr1 141474-173862      - |   HAVANA     gene ENSG00000241860            7  AL627309.5         lncRNA
  ENSG00000237491     chr1 778747-810065      + |   HAVANA     gene ENSG00000237491           10   LINC01409         lncRNA
  ENSG00000228794     chr1 825138-868202      + |   HAVANA     gene ENSG00000228794            9   LINC01128         lncRNA
  ENSG00000187634     chr1 923928-944581      + |   HAVANA     gene ENSG00000187634           12      SAMD11 protein_coding
  ENSG00000188976     chr1 944203-959309      - |   HAVANA     gene ENSG00000188976           11       NOC2L protein_coding
  ENSG00000187961     chr1 960584-965719      + |   HAVANA     gene ENSG00000187961           14      KLHL17 protein_coding



```{r merge rowRanges across all datasets}
flist <- list(neuropil="pseudo_neuropil_pos_donor_spd.rds",
                neun="pseudo_neun_pos_donor_spd.rds",
                pnn="pseudo_pnn_pos_donor_spd.rds",
                vasc="pseudo_vasc_pos_donor_spd.rds" ##-- missing slide_id
              )
dlist <- lapply(flist, function(x) {
      spe <- readRDS(file.path(datadir, "PB_dx_spg", x))
      agg <- scuttle::aggregateAcrossCells(spe, DataFrame(donor = spe$brnum), store_number = "n_SpDs")
      ## colnames are lost during aggregation, restore them from donor
      ncells_sum <- rowsum(
        x = matrix(spe$ncells, ncol = 1),
        group = spe$brnum,  na.rm = TRUE
      )
      donor_ids <- colData(agg)$donor
      colData(agg)$ncells <- ncells_sum[donor_ids, 1]
      agg
  })
dlist$spds <- d
```

```{r merge and save gene info}
all_granges <- lapply(dlist, rowRanges)
merged_granges <- do.call(c, unname(all_granges))
merged_granges <- merged_granges[!duplicated(names(merged_granges))]
fpath <- file.path(refdir, "granges.qs2")
if (!file.exists(fpath)) {
  qs_save(merged_granges, file=fpath)
}
```