---
title: "eqtl results exploration"
##output: html_notebook
---
```{r}
suppressMessages({
    library(arrow)
    library(data.table)
    library(BiocParallel)
    library(qs2)
    library(SummarizedExperiment)
    library(stringi)
    library(GenomicRanges)
    library(here)
  })
here::i_am('.git/HEAD')
rdsdir=here("processed-data", "rds")
## refdir-> ref folder symlink to CHESS Brain ref data for now
refdir=here("processed-data", "ref")
```


```{r}
## load all rowRanges
granges <- qs_read(file.path(refdir, "granges.qs2")) ## gene info file created by 00_explore_data.Rmd last chunk
g2sym <- setNames(granges$gene_name, granges$gene_id)

fgwasflt <- file.path(refdir, 'SCZ_GWAS_flt_1e-6_R.8.tab.gz')
if (file.exists(fgwasflt)) {
    gwas <- fread(fgwasflt) # 30205 variants
} else {
  stop("file not found: ", fgwasflt) ## are you sure you want to rebuild this?
  allgwas <- fread(file.path(refdir, 'SCZ_GWAS_hg38_euro.tab.gz')) # 7659767
  ##GWAS human standard: < 5e-8
  ##         suggestive: < 1e-6
  gwas <- allgwas[p<1e-6 & !is.na(pos) & impinfo>=0.8] ## 30205 variants
  fwrite(gwas, fgwasflt, sep='\t', quote=FALSE)
}
gwas_set <- unique(gwas[p<5e-8]$variant_id) ## GWAS standard risk threshold
gwas1e6_set <- unique(gwas$variant_id) ## loose p < 1e-6
## silly ambiguity mismatch possible between GWAS IDs and genotype data, so add the swapped bases versions
swap_alleles <- function(variant_id) {
   parts <- strsplit(variant_id, ":")
   sapply(parts, function(p) {
     if (length(p) == 4) {
       paste(p[1], p[2], p[4], p[3], sep=":")
     } else {
       variant_id
     }
   })
 }
# Create extended set with both original and swapped IDs
gwas_set_swapped <- swap_alleles(gwas_set)
gwas_set_ext <- unique(c(gwas_set, gwas_set_swapped))

```

```{r gwas-tqtl overlap prep}
## not directly used below, but in the 03b_eQTL_boxplots.Rmd
## tensorqtl variant universe:
pvar <- fread(here("processed-data/genotypes/plink2", "merged_maf05.pvar"), sep = "\t")
setnames(pvar, '#CHROM', 'CHROM')

## tensorqtl used PLINK2 .pvar ID for variants: chr:pos:REF:ALT
pvar_info <- pvar[, .(chr = as.character(CHROM),
  pos = as.integer(POS),
  a0  = REF, a1  = ALT, ID)]
## MAPK3 eQTL: "chr16:30114519:G:C" = rs7542
fgwas <- file.path(refdir, 'GWAS_SCZ_tqtl-matched.tab.gz') ## rebuild if not found
if (file.exists(fgwas)) {
  gwas_tqtl <- fread(fgwas)
} else {
  library(bigsnpr) # remotes::install_github("privefl/bigsnpr")
  library(bigreadr) ## for fread2
  fgwas38 <- file.path(refdir, 'scz_gwas_wide_hg38.qs2')
  if (file.exists(fgwas38)) {
    gwas_wide <- qs_read(fgwas38)
  } else {
    fhg19gwas <- file.path(refdir, 'PGC3_SCZ_wave3.european.autosome.public.v3.vcf.tsv.gz')
    ## NOTE: A1 as the allele for frequency/beta so BETA is "of A1".
    ##    A1/A2 are not guaranteed to be genome REF/ALT. They're effect-allele oriented (A1) and the "other" allele (A2)

    #data/PGC3_SCZ_wave3.european.autosome.public.v3.vcf.tsv.gz'
    stopifnot(file.exists(fhg19gwas))
    liftOver <- Sys.which("liftOver")
    stopifnot(file.exists(liftOver))

    gwas_wide <- fread2(fhg19gwas)
    colnames(gwas_wide) <- tolower(colnames(gwas_wide))
    setnames(gwas_wide, old = c("chrom","id","a1","a2","se",    "neff","pval"),
                        new = c("chr","rsid","a0","a1","beta_se","N",  "p"))
    ## in this notation, a0 is the EFFECT allele (former A1) and a1 is the other allele (former A2)
    setcolorder(gwas_wide, c("rsid","chr","pos","a0","a1","beta","beta_se","N","p"))
    gwas_wide$chr <- as.character(gwas_wide$chr)

    gwas_wide <- snp_modifyBuild(gwas_wide, liftOver, from = 'hg19', to = 'hg38')
    ## 4967 variants have not been mapped.
    library(SNPlocs.Hsapiens.dbSNP155.GRCh38)
    dbsnps <- SNPlocs.Hsapiens.dbSNP155.GRCh38
    setDT(gwas_wide)
    gwas_wide$chr <- as.character(gwas_wide$chr)
    ## only pull the unmapped unique rsids:
    ids_unmapped <- unique(gwas_wide[is.na(pos) & grepl("^rs\\d+$", rsid), rsid])
    ### 4899 ids to map
    ## annoyingly slow:
    rehg38 <- snpsById(dbsnps, ids=ids_unmapped, ifnotfound="drop")
    ## rescued 4360 variants
    ## Normalize the SNPlocs result to chr/pos and join-update only rows with missing pos
    hdt <- as.data.table(rehg38)
    setnames(hdt, c('seqnames'), c('chr'))
    hdt$chr <- as.character(hdt$chr)
    setkey(hdt, RefSNP_id)
    setkey(gwas_wide, rsid)
    gwas_wide[hdt$RefSNP_id, `:=`(chr=hdt$chr, pos=hdt$pos)]
     ## only 607 variants are not mapped
    gwas_wide[, variant_id := sprintf("chr%s:%s:%s:%s", chr, pos, a0, a1)]
    gwas_wide[, chr := paste0("chr", chr)]
    qs_save(gwas_wide, fgwas38)
  }
  ### NOTE: in gwas_wide, a0 = EFFECT allele (former A1), a1 = other allele (former A2)
  ## harmonize with tensorqtl variant universe
  ## prepare GWAS for matching
  sumstats_for_match <- gwas_wide[, .(
    chr, pos,
    a0 = a1,      # swap alleles to match bigsnpr notation (a0 = former A2 = the OTHER allele)
    a1 = a0,      # a1 = former A1 = the EFFECT allele
    beta, beta_se, N, p, ncas, ncon, impinfo, fcas, fcon, rsid
  )]
  ## bigsnpr removes palindromic A/T and C/G when strand_flip=TRUE; match those separately
  is_ambiguous <- with(sumstats_for_match, paste(a0, a1) %in% c("A T", "T A", "C G", "G C"))
  sumstats_nonambig <- sumstats_for_match[!is_ambiguous]
  sumstats_ambig    <- sumstats_for_match[is_ambiguous]
  gwas_m_nonambig <- if (nrow(sumstats_nonambig)) {
    bigsnpr::snp_match(
      sumstats_nonambig, pvar_info,
      strand_flip = TRUE, return_flip_and_rev = TRUE
    )
    # 6,500,465 variants to be matched.
    # 0 ambiguous SNPs have been removed.
    # 4,390,082 variants have been matched; 4,057 were flipped and 3,287,160 were reversed.
  }
  ## For palindromic SNPs, only allow same-strand matching (no strand flip)
  gwas_m_ambig <- if (nrow(sumstats_ambig)) {
    bigsnpr::snp_match(
      sumstats_ambig, pvar_info,
      strand_flip = FALSE, return_flip_and_rev = TRUE
    )
    # 1,159,302 variants to be matched.
    # 780,884 variants have been matched; 0 were flipped and 585,387 were reversed.
  }
  gwas_m <- data.table::rbindlist(
    list(gwas_m_nonambig, gwas_m_ambig),
    use.names = TRUE, fill = TRUE
  )
  ## 5170966 rows total
  # gwas_m now has alleles aligned to pvar_info, and beta flipped if alleles swapped
  setDT(gwas_m)
  # Rebuild variant_id to match tensorQTL
  gwas_m[, variant_id := pvar_info$ID[`_NUM_ID_`]]

  gwas_tqtl <- gwas_m[, .(
    variant_id, A1=a1, A2=a0, ## A1 is the EFFECT allele!
    beta, beta_se,
    N,  p,  ncas, ncon, impinfo,
    rsid, fcas, fcon
  )]
  ## deduplicate defensively
  setkey(gwas_tqtl, variant_id)
  gwas_tqtl <- unique(gwas_tqtl)
  ##
  fwrite(gwas_tqtl, fgwas, sep = "\t")
}

```


```{r load tensor QTL results except nominal}
pats <- c('\\.map_cis\\.tab\\.gz$', '\\.map_independent\\.txt\\.gz$',
           '\\.DxINT\\.cis_qtl_top_assoc\\.txt\\.gz$')
npats <- c('map', 'indep', 'dxint')
cisres <- list()
for (i in seq_along(pats)) {
  fcis <- list.files(path='tqtl_out', pattern=pats[i], recursive = TRUE, full.names = TRUE)
  names(fcis) <- sub("/", "", stri_match_first_regex(fcis, '/(\\w+)\\.gene')[,2])
  cisres[[npats[i]]] <- lapply(fcis, fread)
}
```

```{r load DEGS}
## WARNING: makes sure char not factor is passed to this char2char mapping!
ds2name <- setNames( c("SpD01-WMtz", "SpD02-L3/4", "SpD03-L6",  "SpD04-WM", "SpD05-L5",
       "SpD06-L2/3","SpD07-L1", "Neun", "Vasc", "PNN", "Neuropil"),  c(paste0('spd0',1:7), "neun", "vasc", "pnn", "neuropil"))

## load degs adust SpD as SpD degs
useLayerAdjusted=FALSE
ds2degs <- list()
if (useLayerAdjusted) {
  deg_spd_adj <- fread(file.path(rdsdir, '10_dx_deg_adjust_spd/dx-deg_PRECAST07.csv'), data.table = FALSE)
  rownames(deg_spd_adj) <- deg_spd_adj$V1
  setnames(deg_spd_adj, 'V1', 'fid')
  spd_degs <- subset(deg_spd_adj, fdr_scz<=0.1) ## 10% FDR threshold was used
  spds <- grep('spd', names(ds2name), value=TRUE)
  ds2degs <- lapply(spds, \(x) spd_degs)
  names(ds2degs) <- spds
} else { # use layer-restricted
  for (ds in grep("^spd", names(ds2name), value=TRUE)) {
    degspd <- fread(file.path(rdsdir, "11_dx_deg_interaction", sprintf("layer_restricted_logFC_%s.csv", ds)),
                     data.table = FALSE)
    degspd <- subset(degspd, adj.P.Val<0.1)
    degspd$fid <- degspd$gene_id
    rownames(degspd) <- degspd$gene_id
    setcolorder(degspd, 'fid')
    ds2degs[[ds]] <- degspd
  }
}

## for the SpG contexts, separate DEGs per context:
spgs <- grep('^[^s]', names(ds2name), value=TRUE)
for (ds in spgs) {
   if (ds=="vasc") {
     fdegs <- here("processed-data", "spg_pb_de", "test_SPD_pseudo_vasc_pos.csv")
   } else {
     fdegs <- here("code/analysis/", paste0("dx_deg_spg_", ds), paste0(ds,"-dx_DEG-GM.csv"))
   }
   if (!file.exists(fdegs)) {
     stop("file not found: ", fdegs)
   }
   dt <- fread(fdegs, data.table = FALSE)
   rownames(dt) <- dt$ensembl
   dt$fid <- rownames(dt)
   setcolorder(dt, 'fid')
   dt <- subset(dt, fdr_scz<=0.1)
   ds2degs[[ds]] <- dt
}
# pnn has no degs for FDR<0.1 ?
```

```{r}
## tag dataset to each data.table before merging
tag_ll <- function(dt_list) {
  rbindlist(lapply(names(dt_list), function(nm) {
    ctx_degs <- ds2degs[[nm]]
    x <- copy(dt_list[[nm]])
    x[, ds:=nm]
    x[, gid := phenotype_id]
    ## add gtype and symbol
    x[, symbol :=g2sym[phenotype_id]]
    x[, ds := as.character(ds)]
    x[, dge:=as.integer(phenotype_id %chin% ctx_degs$fid)]
    x[, gwas := as.integer(variant_id %chin% gwas_set_ext)]
    x
  }))
}

map_all   <- tag_ll(cisres$map)     # has qval
indep_all <- tag_ll(cisres$indep)   # has rank
dx_all    <- tag_ll(cisres$dxint)   # has pval_gi, b_g, b_gi, pval_adj_bh

#dedup_by_ds <- function(dt) unique(dt, by = c("ds","phenotype_id"))

```

```{r discovery and analysis}
get_qmap  <- function(emap, q) {
  emap[qval < q, .(
    n_eGenes = uniqueN(phenotype_id),                          # unique eGenes
    n_GWAS      = uniqueN(phenotype_id[gwas == 1]),          # eGenes whose lead is in GWAS (5e-8)
    GWAS = paste(sort(unique(symbol[gwas==1])), collapse=", "),
    n_DEG   = uniqueN(gid[dge   == 1]),
    DEGS = paste(sort(unique(symbol[dge==1])), collapse=", "),
    n_DEG_GWAS =  uniqueN(phenotype_id[dge==1 & gwas==1]),
    DEG_GWAS = paste(sort(unique(symbol[dge==1 & gwas==1])), collapse=", ")
  ),
  by = .(ds)
][order(ds)]
}

## eGene cis_map counts at eQTL FDR < 5%
mapsum05 <- get_qmap(map_all, 0.05)

## reorder rows by matching ds to the order in names(ds2name)
mapsum05 <- mapsum05[match(names(ds2name), mapsum05$ds), ]
## Now replace with nicer names
mapsum05$ds <- ds2name[mapsum05$ds]
## show and save the table
mapsum05
fwrite(mapsum05, "map_cis_summary.csv")
```

            ds n_eGenes n_GWAS                           GWAS n_DEG                                                                       DEGS n_DEG_GWAS   DEG_GWAS
        <char>    <int>  <int>                         <char> <int>                                                                     <char>      <int>     <char>
 1:       Neun      134      2              LRRC37A2, ZSCAN31     7          GABPB1-AS1, GATD3A, KANSL1-AS1, LINC01535, NIPBL-DT, RPS26, XRRA1          0
 2:   Neuropil      161      4 HLA-DMA, MAPK3, VARS2, ZSCAN31     9 GATD3A, KANSL1-AS1, LINC01535, MAPK3, NIPBL-DT, NUDT14, RPS26, SVBP, XRRA1          1      MAPK3
 3:        PNN        5      0                                    1                                                                      RPS26          0
 4: SpD01-WMtz       10      0                                    1                                                                      RPS26          0
 5: SpD02-L3/4      179      2              LRRC37A2, ZSCAN31     8  GABPB1-AS1, GATD3A, KANSL1-AS1, LINC01535, NIPBL-DT, NUDT14, RPS26, XRRA1          0
 6:   SpD03-L6       48      2                HLA-C, LRRC37A2     7          GABPB1-AS1, GATD3A, KANSL1-AS1, LINC01535, NIPBL-DT, RPS26, XRRA1          0
 7:   SpD04-WM        2      0                                    1                                                                      RPS26          0
 8:   SpD05-L5       61      1                       LRRC37A2     4                                         GATD3A, LINC01535, NIPBL-DT, RPS26          0
 9: SpD06-L2/3       61      2           KANSL1-AS1, LRRC37A2     6                      GATD3A, KANSL1-AS1, LINC01535, NIPBL-DT, RPS26, XRRA1          1 KANSL1-AS1
10:   SpD07-L1       12      1                     KANSL1-AS1     3                                                  GATD3A, KANSL1-AS1, RPS26          1 KANSL1-AS1
11:       Vasc        3      0                                    1                                                                      RPS26          0


```{r plot barchart}
library(ggplot2)

eqtl_barchart  <- function(mapsum, indep=FALSE, fill=NULL) {
  # pal_lib <- c(PolyA = "#82BEEE", RiboZ = "#FFBE6A")
  dt <- copy(mapsum)
  dt[, context  := as.factor(ds)] ## set levels explictly here
  dt[, lab_above := sprintf("GWAS: %d\nDEG: %d\nDEG∩GWAS: %d", n_GWAS, n_DEG, n_DEG_GWAS)]
  if (indep) {
    ytitle <- "Total independent cis-eQTL signals (p_perm < 0.05)"
    title  <- "Independent cis-eQTLs by context"
  } else {
    ytitle <- "eGenes discovered (lead cis-eQTLs, q < 0.05)"
    title  <- "Lead cis-eQTLs per eGene by context"
  }
  #mapsum[, lab_count := sprintf("%d", n_eGenes)]
  min_count <- min(dt[n_eGenes>0]$n_eGenes)
  # label positions inside each bar (relative to its own height)
  dt[, y_top  := n_eGenes]                    # top-of-bar (for the big count)
  dt[, y_gwas := n_eGenes - min_count * 0.08]             # just below the top
  dt[, y_de   := n_eGenes - min_count * 0.08]             # a bit further down, top-aligned

  # two rows per bar: GWAS and Other; stack within each bar
  bar_df <- rbindlist(list(
    dt[, .(context, category="GWAS (p<5e-8)", count=n_GWAS)],
    dt[, .(context, category="Other",         count=n_eGenes - n_GWAS)]
  ))
  if (is.null(fill)) {
     p <- ggplot(dt, aes(x=context, y=n_eGenes, fill=context)) +
       geom_col(width=0.7, color="grey30", linewidth=0.2)
  } else {
     p <- ggplot(dt, aes(x=context, y=n_eGenes)) +
       geom_col(width=0.7, color="grey30", linewidth=0.2, fill=fill)
  }
  p <- p +
    #facet_grid(level ~ coding, scales="free_y") +
    # big total on top of each bar
    geom_text(aes(y = y_top, label = n_eGenes), vjust = -0.3,
              fontface = "bold", size = 3.3, color = "black") +
    # inside-bar annotations
    geom_text(aes(y = y_top, label = lab_above), vjust = -0.5,
              color = "black", size = 3) +
    # scale_fill_manual(values = pal_lib, guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.20))) +
    coord_cartesian(clip = "off") +
    labs(x = NULL, y = ytitle, title = title) +
    theme_bw(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      axis.ticks.x     = element_blank(),
      strip.background = element_rect(fill = "grey95", color = "grey80")
    )
 p
}
print(eqtl_barchart(mapsum05[grepl('^SpD', ds)]))
print(eqtl_barchart(mapsum05[!grepl('^SpD', ds)], fill="#A0A0A0"))
```

```{r independent signals}
## Add gwas1e6 column to indep_all ?
#indep_all[, gwas1e6 := as.integer(variant_id %chin% gwas1e6_set)]

## independent signals per feature (rank is 1..k per phenotype)

alpha <- 0.05  # primary FDR line for discovery - typically 0.05
## parent eFeature q-values from map_all
parent_q <- map_all[, .(phenotype_id, ds, qval_parent = qval)]

## Note that independent signals are already filtered by pval_beta<=0.05 in the parent eFeature
## since cis.map_independent() only tests features with pval_beta<=0.05 from cis.map_cis
### -- join and gate using pval_perm, which is slightly more conservative than pval_beta
indep_f <- merge(indep_all, parent_q, by=c("phenotype_id","ds"))
indep_f <- indep_f[qval_parent < alpha & pval_perm < 0.05]
#indep_f[, gwas1e6 := as.integer(variant_id %chin% gwas1e6_set)]

## Per-feature summary
#indep_feat <- indep_f[, .(
#  n_signals   = max(rank),
#  any_gwas    = as.integer(any(gwas == 1)),
#  any_gwas1e6 = as.integer(any(gwas1e6 == 1)),
#  any_dge     = as.integer(any(dge == 1))
#), by = .(ds, phenotype_id, gid, symbol)]

## Summary for plotting (counting total independent signals)
indep_summary <- indep_f[, .(
  n_eGenes = .N,  # total independent signals
  n_GWAS = sum(gwas),
  n_DEG = uniqueN(gid[dge == 1]),
  n_DEG_GWAS = sum(dge == 1 & gwas == 1)
), by = ds][order(ds)]


## reorder rows by matching ds to the order in names(ds2name)
indep_summary <- indep_summary[match(names(ds2name), indep_summary$ds), ]


indep_summary$ds <- ds2name[indep_summary$ds]

print(eqtl_barchart(indep_summary[grepl('^SpD', ds)], indep=TRUE))
print(eqtl_barchart(indep_summary[!grepl('^SpD', ds)], fill="#A0A0A0", indep=TRUE))

```


```{r interaction summary}
## choose thresholds
#thr_primary <- 0.10  # primary significance threshold
thr_primary <- 0.25  # exploratory
dx_dups <- dx_all[, .N, by=.(ds, phenotype_id)][N > 1]
if (nrow(dx_dups) > 0) {
  ## there are duplicate entries in dx_all for some features; keeping the first only
  #print(dx_dups)
  dx_all <- dx_all[order(pval_gi)][, .SD[1], by=.(ds, phenotype_id)]
}
dx_best_qc <- dx_all[af >= 0.05 & af <= 0.95 & ma_count >= 20]
## primary and exploratory thresholds
dx_primary  <- dx_best_qc[pval_adj_bh <= thr_primary]
#dx_explore  <- dx_best_qc[pval_adj_bh <= 0.25]

summ_dx <- function(dt) dt[, .(
  n_feat        = .N,
  prop_pos_bgi  = mean(b_gi > 0),
  med_abs_bgi   = median(abs(b_gi)),
  n_GWAS        = sum(gwas),
  n_DEG    = uniqueN(gid[dge == 1]),
), by=.(ds)][order(ds)]

dxsum10 <- summ_dx(dx_primary)

```
