---
title: "eQTL genotype boxplots"
##output: html_notebook
---
```{r}
suppressMessages({
    library(data.table)
    library(BiocParallel)
    library(qs2)
    library(SummarizedExperiment)
    library(stringi)
    library(GenomicRanges)
    library(ggplot2)
    library(pgenlibr)
    library(here)
  })
here::i_am('.git/HEAD')
rdsdir=here("processed-data", "rds")
## refdata -> ref folder symlink to CHESS Brain ref data for now
refdir=here("processed-data", "ref") ## some general reference files prepared for our data
tqtl_out=here("processed-data", "eQTL/tqtl_out")
tqtl_in=here("processed-data", "eQTL/tqtl_in")

ds2name <- setNames( c("SpD01-WMtz", "SpD02-L3/4", "SpD03-L6",  "SpD04-WM", "SpD05-L5",
       "SpD06-L2/3","SpD07-L1", "Neun", "Vasc", "PNN", "Neuropil"),  c(paste0('spd0',1:7), "neun", "vasc", "pnn", "neuropil"))
dss <- names(ds2name)
```

```{r load our gwas-matched data}

fgwas <- file.path(refdir, 'GWAS_SCZ_tqtl-matched.tab.gz') ## this was created in 03_eqtl_explore.Rmd
if (file.exists(fgwas)) {
  gwas_vars <- fread(fgwas)
} else {
  stop("GWAS-tensorQTL variant-harmonized file not found: ", fgwas, "\nRun 03_eqtl_explore.Rmd to create it.")
}

## load gene info
granges <- qs_read(file.path(refdir,"granges.qs2")) ## gene info file created by 00_explore_data.Rmd last chunk
g2sym <- setNames(granges$gene_name, granges$gene_id)

```

NOTE: for GWAS SCZ data, A1 as the allele for frequency/beta so BETA is "of A1".
Thus, A1/A2 are not guaranteed to be genome REF/ALT.
They're effect-allele oriented (A1), and A2 is just the "other" allele,
either could be REF or ALT, relative to the reference genome.

```{r-console-output}
> head(gwas_vars)
            variant_id     A1     A2         beta beta_se        N      p  ncas  ncon impinfo       rsid  fcas  fcon
                <char> <char> <char>        <num>   <num>    <num>  <num> <int> <int>   <num>     <char> <num> <num>
1: chr10:100000235:C:T      T      C -0.009098483  0.0094 58749.13 0.3310 53386 77258   0.983 rs11596870 0.692 0.692
2: chr10:100000943:G:A      A      G  0.017695649  0.0139 58749.13 0.2015 53386 77258   0.986 rs11190359 0.890 0.893
3: chr10:100000979:T:C      C      T  0.020804929  0.0183 58749.13 0.2559 53386 77258   0.995 rs11190360 0.941 0.942
4: chr10:100002628:A:C      C      A -0.006200736  0.0087 58749.13 0.4760 53386 77258   0.991 rs11190363 0.541 0.546
5: chr10:100002875:A:G      G      A  0.014900463  0.0161 58749.13 0.3574 53386 77258   0.995  rs7894103 0.921 0.924
6: chr10:100004256:T:A      A      T -0.007402534  0.0090 58749.13 0.4113 53386 77258   0.983  rs7915731 0.631 0.633
```


```{r load all independent tensor QTL results}
pat <- '\\.map_independent\\.txt\\.gz$'
fcis <- list.files(path=tqtl_out, pattern=pat, recursive = TRUE, full.names = TRUE)
names(fcis) <- sub("/", "", stri_match_first_regex(fcis, '/(\\w+)\\.gene')[,2])
## names(fcis) must match dss items and we have to sort it to be in the same order
stopifnot("fcis/dss name mismatch" = setequal(names(fcis), dss))
fcis <- fcis[dss] # sort in dss order

indres <- lapply(names(fcis), \(x) {
    dt <- fread(fcis[[x]])
    if (nrow(dt)==0) message("Dataset ", x, " has no rows!")
    dt[, gene_name:=g2sym[phenotype_id]]
    })
names(indres) <- dss
```

```{r load all nominal eQTL results}
pat <- '\\.eqtl-nominal_FDR05\\.tab\\.gz$'
fnom <- list.files(path=tqtl_out, pattern=pat, recursive = TRUE, full.names = TRUE)
names(fnom) <- sub("/", "", stri_match_first_regex(fnom, '/(\\w+)\\.')[,2])
stopifnot("fnom/dss name mismatch" = setequal(names(fnom), dss))
fnom <- fnom[dss] # sort in dss order

rsid_lut <- unique(gwas_vars[, .(variant_id, rsid)])

nomres <- lapply(names(fnom), \(x) {
    dt <- fread(fnom[[x]])
    if (nrow(dt)==0) message("Dataset ", x, " has no rows!")
    dt[, gene_name:=g2sym[phenotype_id]]
    dt <- merge(
      dt,
      rsid_lut,
      by = "variant_id",
      all.x = TRUE,
      sort = FALSE
    )
    dt
    })
names(nomres) <- dss
```


```{r load genotypes}
pgen_prefix <- here("processed-data", "genotypes", "plink2", "merged_maf05")

#' Load and prepare genotype data from plink2 files
#' @param pgen_prefix Path prefix to plink2 files (without .pgen/.pvar/.psam extension)
#' @return List with: dosage (samples x variants matrix),
#'         pvar (data.table with variant info), samples (character vector of sample IDs)
#' if gwas_vars is given, relevant GWAS info will be added to the pvar data.table, where available, columns:
#' rsid, p, A1, A2, beta, beta_se, impinfo
prepGenotypes <- function(pgen_prefix, gwas_vars = NULL) {
  # File paths
  pgen_file <- paste0(pgen_prefix, ".pgen")
  pvar_file <- paste0(pgen_prefix, ".pvar")
  psam_file <- paste0(pgen_prefix, ".psam")
  # Check files exist
  stopifnot("pgen file not found" = file.exists(pgen_file))
  stopifnot("pvar file not found" = file.exists(pvar_file))
  stopifnot("psam file not found" = file.exists(psam_file))
  # Load pvar (variant info)
  pvar <- fread(pvar_file, sep = "\t")
  setnames(pvar, "#CHROM", "CHROM")
  pvar[, var_idx := .I]  # 1-based index

  # Attach GWAS info if provided
  if (!is.null(gwas_vars)) {
    required_cols <- c("variant_id", "rsid", "A1", "A2", "beta", "beta_se", "N", "p", "impinfo")
    stopifnot("gwas_vars must include required columns" = all(required_cols %in% names(gwas_vars)))
    gwas_sub <- as.data.table(gwas_vars)[, ..required_cols]
    pvar <- merge(
      pvar,
      gwas_sub,
      by.x = "ID",
      by.y = "variant_id",
      all.x = TRUE,
      sort = FALSE
    )
  }
  # Add a secondary index for fast ID lookups without reordering rows
  setindexv(pvar, "ID")
  # Load psam (sample info)
  psam <- fread(psam_file, sep = "\t")
  # psam typically has #IID column
  if ("#IID" %in% names(psam)) {
    samples <- psam[["#IID"]]
  } else if ("IID" %in% names(psam)) {
    samples <- psam[["IID"]]
  } else {
    stop("Cannot find sample ID column in psam file")
  }
  n_samples <- length(samples)
  n_variants <- nrow(pvar)
  message(sprintf("Loading %d variants x %d samples...", n_variants, n_samples))
  # Open pgen file
  pgen <- NewPgen(pgen_file)
  # Read all genotypes - ReadList returns samples x variants matrix
  # variant_subset uses 1-based indices
  dosage <- ReadList(pgen, seq_len(n_variants), meanimpute = FALSE)
  # Close pgen
  rm(pgen)
  gc()
  # Set row/col names
  rownames(dosage) <- samples
  colnames(dosage) <- pvar$ID
  message("Done loading genotypes.")

  list(
    dosage = dosage,
    pvar = pvar,
    samples = samples
  )
}
## Load genotypes (8GB RAM needed just for this)
## this prep takes a long time, might be easier to save it in refdir after preparing it once
fprepgeno <- file.path(refdir, "plink_pgen_data_objects.qs2")
if (!file.exists(fprepgeno)) {
  geno <- prepGenotypes(pgen_prefix, gwas_vars)
  qs_save(geno, file=fprepgeno)
} else {
  message("loading prepared genotype data.. please wait")
  ## even loading is slow here
  geno <- qs_read(fprepgeno)
}
```

```{r orientation helpers}
#' Build per-variant orientation metadata versus pvar allele order.
#' @param variant_ids Character vector of variant IDs formatted as chr:pos:REF:ALT
#' @param pvar data.table with ID, REF, ALT columns
#' @return data.table with orientation columns for each variant_id
build_orientation_lut <- function(variant_ids, pvar) {
  ids <- data.table(variant_id = unique(as.character(variant_ids)))
  ids <- ids[!is.na(variant_id) & nzchar(variant_id)]
  ids[, c("id_chr", "id_pos", "id_REF", "id_ALT") := tstrsplit(
    variant_id, ":", fixed = TRUE, keep = 1:4
  )]

  pvar_lut <- pvar[, .(variant_id = ID, pvar_REF = REF, pvar_ALT = ALT)]
  setkey(pvar_lut, variant_id)
  ori <- pvar_lut[ids, on = "variant_id"]

  ori[, orientation_status := fifelse(
    is.na(pvar_REF) | is.na(pvar_ALT), "missing_pvar",
    fifelse(
      is.na(id_REF) | is.na(id_ALT), "mismatch",
      fifelse(
        id_REF == pvar_REF & id_ALT == pvar_ALT, "aligned",
        fifelse(id_REF == pvar_ALT & id_ALT == pvar_REF, "swapped", "mismatch")
      )
    )
  )]
  ori[, orientation_sign := fifelse(
    orientation_status == "aligned", 1L,
    fifelse(orientation_status == "swapped", -1L, NA_integer_)
  )]
  ori[]
}

#' Apply orientation sign to a slope-like effect.
#' @param x Numeric vector
#' @param orientation_sign Integer scalar/vector in {-1,1,NA}
#' @return Numeric vector in reference-genome orientation
orient_effect <- function(x, orientation_sign) {
  ifelse(is.na(orientation_sign), x, x * orientation_sign)
}
```


```{r MAKP3 eQTL data only in neuropil}
## let's focus on 1 independent result for testing:
ds <- 'neuropil'
dsname <- unname(ds2name[ds])
## load data for the specified dataset
spe <- qs_read(file.path(tqtl_in, sprintf('%s.gene.spe.qs2',ds)))
## use assays(spe)$tmm for expression data, TMM normalized log(CPM+1)
covar_file <- file.path(tqtl_in, sprintf("%s.gene.covars.txt", ds))
indr <- indres[[ds]]
stopifnot(!is.null(indr))
## load Xenium donor list:
fxenium <- here("processed-data", "eQTL", "xenium_brnums.keep")
xdonors <- fread(fxenium, header=FALSE)$V1
```


```{r prep_eQTLPlotData}
#' Prepare eQTL plot data by residualizing expression against covariates
#' @param spe SpatialExperiment object with tmm assay
#' @param geno Genotype object from prepGenotypes()
#' @param covar_file Path to tensorQTL covariates file
#' @return List with residualized/raw expression matrices, gene/sample info, and genotype reference
prep_eQTLPlotData <- function(spe, geno, covar_file) {
  # ---- Validate inputs ----
  stopifnot("spe must be a SummarizedExperiment" = is(spe, "SummarizedExperiment"))
  stopifnot("geno must be a list with dosage, pvar, samples" =
              all(c("dosage", "pvar", "samples") %in% names(geno)))
  stopifnot("covar_file must exist" = file.exists(covar_file))
  stopifnot("spe must have 'tmm' assay" = "tmm" %in% assayNames(spe))
  # ---- Load covariates ----
  message("Loading covariates from: ", covar_file)
  covars_raw <- fread(covar_file, sep = "\t")
  covar_names <- covars_raw$id
  covar_samples <- setdiff(names(covars_raw), "id")
  # Transpose: samples as rows, covariates as columns
  covar_mat <- t(as.matrix(covars_raw[, -1, with = FALSE]))
  colnames(covar_mat) <- covar_names
  rownames(covar_mat) <- covar_samples
  # Remove intercept if present (cleaningY doesn't need it)
  if ("(Intercept)" %in% colnames(covar_mat)) {
    covar_mat <- covar_mat[, colnames(covar_mat) != "(Intercept)", drop = FALSE]
  }
  # ---- Align samples across all sources ----
  spe_samples <- colnames(spe)
  geno_samples <- geno$samples
  message(sprintf("Sample counts: SPE=%d, Genotypes=%d, Covariates=%d",
                  length(spe_samples), length(geno_samples), length(covar_samples)))
  # Find common samples
  common_samples <- Reduce(intersect, list(spe_samples, geno_samples, covar_samples))
  if (length(common_samples) == 0) {
    stop("No common samples found across SPE, genotypes, and covariates!")
  }
  n_dropped <- length(spe_samples) - length(common_samples)
  if (n_dropped > 0) {
    warning(sprintf("Dropped %d samples not present in all data sources. Using %d samples.",
                    n_dropped, length(common_samples)))
  }
  message(sprintf("Using %d common samples", length(common_samples)))
  # Reorder all data to common sample order
  spe_idx <- match(common_samples, spe_samples)
  geno_idx <- match(common_samples, geno_samples)
  covar_idx <- match(common_samples, covar_samples)
  # Subset and reorder covariate matrix
  covar_mat <- covar_mat[covar_idx, , drop = FALSE]
  # ---- Extract expression data ----
  message("Extracting expression data...")
  raw_expr <- assays(spe)$tmm[, spe_idx, drop = FALSE]
  colnames(raw_expr) <- common_samples
  # ---- Residualize expression ----
  # cleaningY: P = number of columns to PROTECT (keep) from left of model matrix
  # We want to regress out all covariates to show pure genotype effect
  # P=0 means regress out everything; but cleaningY needs at least the intercept
  # Since we removed intercept, we need to add it back for cleaningY to work properly
  message("Residualizing expression with jaffelab::cleaningY...")
  # Add intercept column as first column (cleaningY needs it)
  mod_mat <- cbind("(Intercept)" = 1, covar_mat)
  # P=1 protects only intercept, regresses out all other covariates
  resid_expr <- jaffelab::cleaningY(raw_expr, mod = mod_mat, P = 1)
  colnames(resid_expr) <- common_samples
  # ---- Build gene info ----
  rr <- rowRanges(spe)
  gene_info <- data.table(
    gene_id = rownames(spe),
    gene_name = rr$gene_name,
    chr = as.character(seqnames(rr)),
    start = start(rr),
    end = end(rr)
  )
  # ---- Build sample info ----
  cd <- colData(spe)[spe_idx, ]
  sample_info <- data.table(
    sample_id = common_samples,
    DX = cd$DX
  )
  # Add any other useful columns if present
  if ("sex" %in% names(cd)) sample_info[, sex := cd$sex]
  if ("age" %in% names(cd)) sample_info[, age := cd$age]
  # ---- Store sample indices for genotype subsetting ----
  message("Done preparing eQTL plot data.")
  list(
    resid_expr = resid_expr,
    raw_expr = raw_expr,
    gene_info = gene_info,
    sample_info = sample_info,
    sample_idx = geno_idx,  # indices into geno$dosage rows
    geno = geno  # reference to genotype object
  )
}

```

```{r SKIP test/validate prep_eQTLPlotData}
# Verify output structure
##
plotdata <- prep_eQTLPlotData(spe, geno, covar_file)
cat("\n=== prep_eQTLPlotData output ===\n")
cat("resid_expr:", class(plotdata$resid_expr), "dim:", paste(dim(plotdata$resid_expr), collapse=" x "), "\n")
cat("raw_expr:", class(plotdata$raw_expr), "dim:", paste(dim(plotdata$raw_expr), collapse=" x "), "\n")
cat("gene_info:", class(plotdata$gene_info), "nrow:", nrow(plotdata$gene_info), "\n")
cat("sample_info:", class(plotdata$sample_info), "nrow:", nrow(plotdata$sample_info), "\n")
cat("DX distribution:", table(plotdata$sample_info$DX), "\n")

# Test genotype retrieval for an eQTL pair
cat("\n=== Genotype retrieval test ===\n")
test_var <- indr$variant_id[1]
test_gene <- indr$phenotype_id[1]
cat("Test variant:", test_var, "\nTest gene:", test_gene, "\n")

test_ref <- plotdata$geno$pvar$REF[match(test_var, plotdata$geno$pvar$ID)]
test_alt <- plotdata$geno$pvar$ALT[match(test_var, plotdata$geno$pvar$ID)]
test_dosage <- plotdata$geno$dosage[plotdata$sample_idx, test_var]
test_geno_int <- round(test_dosage)
test_geno_labels <- rep(NA_character_, length(test_geno_int))
test_ok <- !is.na(test_geno_int) & test_geno_int %in% 0:2
if (is.na(test_ref) || is.na(test_alt) || !nzchar(test_ref) || !nzchar(test_alt)) {
  test_geno_labels[test_ok] <- as.character(test_geno_int[test_ok])
} else {
  test_labels_map <- c(paste0(test_ref, test_ref), paste0(test_ref, test_alt), paste0(test_alt, test_alt))
  test_geno_labels[test_ok] <- test_labels_map[test_geno_int[test_ok] + 1L]
}
expr_resid <- plotdata$resid_expr[test_gene, ]

cat("\nExpression by genotype (residualized):\n")
for (g in sort(unique(na.omit(test_geno_labels)))) {
  vals <- expr_resid[test_geno_labels == g]
  cat(sprintf("  %s (n=%d): mean=%.3f, sd=%.3f\n", g, length(vals), mean(vals), sd(vals)))
}
cat("\nExpected beta from eQTL:", round(indr$slope[1], 3), "\n")
```


variant_id (genomic) REF/ALT vs GWAS effect allele A1:
Across all 5,170,966 GWAS-matched variants:
 - A1 == variant_id ALT: 3,872,542 (74.90%)
 - A1 == variant_id REF: 1,298,424 (25.10%)

Among GWAS risk variants (p < 5e-8, n=18,853):
 - A1 == variant_id ALT: 14,282 (75.75%)
 - A1 == variant_id REF:  4,571 (24.25%)

A1 == pvar ALT (100%)

```{r plot_eQTL_boxplots}
#' Create faceted boxplots for eQTL gene-variant pairs
#' @param plotdata Prepared data from prep_eQTLPlotData()
#' @param eqtl_pairs data.table/data.frame with columns: phenotype_id, variant_id,
#'        and optionally: slope (beta), pval_nominal, gene_name
#' @param restrict_donors Optional vector of donor IDs to keep
#' @param title Plot title; default "auto" uses "<context> <gene_name> expression".
#'        Set to NULL or "" to suppress title.
#' @param gwas_eff Logical; if TRUE, orient plot to GWAS beta
#'        with GWAS A1/A2/beta in plotdata$geno$pvar (x=0 protective homozygote, x=2 risk homozygote);
#'        otherwise use variant_id REF/ALT orientation (REF homozygote at 0, ALT homozygote at 2)
#' @param use_raw Logical; if TRUE use raw expression, otherwise use residualized (default FALSE)
#' @param ds Optional dataset context key/string used in auto title; if found in ds2name, full context name is used.
#' @return ggplot object
plot_eQTL_boxplots <- function(plotdata, eqtl_pairs, use_raw = FALSE, restrict_donors = NULL, title = "auto", gwas_eff = FALSE, ds = NULL) {
  # ---- Validate inputs ----
  required_cols <- c("phenotype_id", "variant_id")
  stopifnot("eqtl_pairs must have phenotype_id and variant_id columns" =
              all(required_cols %in% names(eqtl_pairs)))
  # Ensure data.table
  if (!is.data.table(eqtl_pairs)) eqtl_pairs <- as.data.table(eqtl_pairs)
  n_pairs <- nrow(eqtl_pairs)
  if (n_pairs == 0) stop("No eQTL pairs provided")
  # ---- Check variants exist in genotype data ----
  missing_vars <- setdiff(eqtl_pairs$variant_id, colnames(plotdata$geno$dosage))
  if (length(missing_vars) > 0) {
    warning(sprintf("Skipping %d variants not found in genotype data: %s",
                    length(missing_vars), paste(head(missing_vars, 3), collapse=", ")))
    eqtl_pairs <- eqtl_pairs[!variant_id %in% missing_vars]
  }
  # ---- Check genes exist in expression data ----
  missing_genes <- setdiff(eqtl_pairs$phenotype_id, rownames(plotdata$resid_expr))
  if (length(missing_genes) > 0) {
    warning(sprintf("Skipping %d genes not found in expression data: %s",
                    length(missing_genes), paste(head(missing_genes, 3), collapse=", ")))
    eqtl_pairs <- eqtl_pairs[!phenotype_id %in% missing_genes]
  }
  if (nrow(eqtl_pairs) == 0) stop("No valid eQTL pairs remain after filtering")
  # ---- Add gene names if not present ----MAPK3 eQTL in Neuropil (FDR (p_perm) = 0.0006
  if (!"gene_name" %in% names(eqtl_pairs)) {
    gene_map <- plotdata$gene_info[, .(gene_id, gene_name)]
    eqtl_pairs <- merge(eqtl_pairs, gene_map, by.x = "phenotype_id", by.y = "gene_id", all.x = TRUE)
    eqtl_pairs[is.na(gene_name), gene_name := phenotype_id]
  }
  # ---- Build plot data ----
  expr_mat <- if (use_raw) plotdata$raw_expr else plotdata$resid_expr
  sample_info <- plotdata$sample_info
  sample_idx <- plotdata$sample_idx
  if (!is.null(restrict_donors)) {
    if (!is.atomic(restrict_donors)) {
      stop("restrict_donors must be a simple vector of donor IDs")
    }
    restrict_donors <- as.character(restrict_donors)
    restrict_donors <- restrict_donors[!is.na(restrict_donors) & nzchar(restrict_donors)]
    keep_samples <- colnames(expr_mat)
    keep <- keep_samples %in% restrict_donors
    if (!any(keep)) {
      stop("No samples left after applying restrict_donors")
    }
    expr_mat <- expr_mat[, keep, drop = FALSE]
    sample_info <- sample_info[sample_info$sample_id %in% keep_samples[keep], ]
    sample_idx <- sample_idx[keep]
  }
  # Build variant -> rsID map if available
  pvar <- plotdata$geno$pvar
  rsid_col <- if ("rsID" %in% names(pvar)) "rsID" else if ("rsid" %in% names(pvar)) "rsid" else NULL
  rsid_map <- NULL
  if (!is.null(rsid_col) && "ID" %in% names(pvar)) {
    rsid_map <- pvar[, .(variant_id = ID, rsid = get(rsid_col))]
  }
  a1_col <- if ("A1" %in% names(pvar)) "A1" else if ("a1" %in% names(pvar)) "a1" else NULL
  a2_col <- if ("A2" %in% names(pvar)) "A2" else if ("a2" %in% names(pvar)) "a2" else NULL
  beta_col <- if ("beta" %in% names(pvar)) "beta" else if ("BETA" %in% names(pvar)) "BETA" else NULL
  gwas_map <- NULL
  if (!is.null(a1_col) && !is.null(a2_col) && !is.null(beta_col) && "ID" %in% names(pvar)) {
    gwas_map <- pvar[, .(
      variant_id = ID,
      gwas_A1 = as.character(get(a1_col)),
      gwas_A2 = as.character(get(a2_col)),
      gwas_beta = suppressWarnings(as.numeric(get(beta_col)))
    )]
    setkey(gwas_map, variant_id)
  }
  # Build orientation lookup table for current variants against pvar allele order
  orientation_lut <- build_orientation_lut(eqtl_pairs$variant_id, pvar)
  setkey(orientation_lut, variant_id)
  make_geno_labels <- function(dosage_vec, ref, alt) {
    geno_int <- round(dosage_vec)
    labels <- rep(NA_character_, length(geno_int))
    ok <- !is.na(geno_int) & geno_int %in% 0:2
    if (is.na(ref) || is.na(alt) || !nzchar(ref) || !nzchar(alt)) {
      labels[ok] <- as.character(geno_int[ok])
      return(labels)
    }
    map <- c(paste0(ref, ref), paste0(ref, alt), paste0(alt, alt))
    labels[ok] <- map[geno_int[ok] + 1L]
    labels
  }
  # Collect data for all pairs
  plot_list <- lapply(seq_len(nrow(eqtl_pairs)), function(i) {
    gene_id <- eqtl_pairs$phenotype_id[i]
    var_id <- eqtl_pairs$variant_id[i]
    gene_name <- eqtl_pairs$gene_name[i]
    # Get rsID if available
    rsid_val <- NA_character_
    if (!is.null(rsid_map)) {
      rsid_val <- rsid_map$rsid[match(var_id, rsid_map$variant_id)]
    }
    # Get expression and genotypes
    expr_vals <- expr_mat[gene_id, ]
    geno_dosage_raw <- plotdata$geno$dosage[sample_idx, var_id]

    orow <- orientation_lut[.(var_id)]
    id_ref <- if (nrow(orow) > 0L) orow$id_REF[1] else NA_character_
    id_alt <- if (nrow(orow) > 0L) orow$id_ALT[1] else NA_character_
    pvar_ref <- if (nrow(orow) > 0L) orow$pvar_REF[1] else NA_character_
    pvar_alt <- if (nrow(orow) > 0L) orow$pvar_ALT[1] else NA_character_
    ori_status <- if (nrow(orow) > 0L) orow$orientation_status[1] else "missing_pvar"
    ori_sign <- if (nrow(orow) > 0L) orow$orientation_sign[1] else NA_integer_
    gwas_row <- if (!is.null(gwas_map)) gwas_map[.(var_id)] else data.table()
    gwas_a1 <- if (nrow(gwas_row) > 0L) gwas_row$gwas_A1[1] else NA_character_
    gwas_a2 <- if (nrow(gwas_row) > 0L) gwas_row$gwas_A2[1] else NA_character_
    gwas_beta <- if (nrow(gwas_row) > 0L) gwas_row$gwas_beta[1] else NA_real_

    ## determine gwas risk-sign orientation against pvar ref/alt
    risk_sign <- NA_integer_
    gwas_status <- "not_requested"
    protective_allele <- NA_character_
    risk_allele <- NA_character_
    if (isTRUE(gwas_eff)) {
      if (is.null(gwas_map)) {
        gwas_status <- "missing_gwas_cols"
      } else if (is.na(pvar_ref) || is.na(pvar_alt) || !nzchar(pvar_ref) || !nzchar(pvar_alt)) {
        gwas_status <- "missing_pvar"
      } else if (is.na(gwas_a1) || is.na(gwas_a2) || !nzchar(gwas_a1) || !nzchar(gwas_a2)) {
        gwas_status <- "missing_gwas"
      } else if (is.na(gwas_beta)) {
        gwas_status <- "missing_beta"
      } else if (gwas_beta == 0) {
        gwas_status <- "zero_beta"
      } else {
        pvar_ref_u <- toupper(pvar_ref)
        pvar_alt_u <- toupper(pvar_alt)
        gwas_a1_u <- toupper(gwas_a1)
        gwas_a2_u <- toupper(gwas_a2)
        if (!((gwas_a1_u == pvar_alt_u && gwas_a2_u == pvar_ref_u) ||
              (gwas_a1_u == pvar_ref_u && gwas_a2_u == pvar_alt_u))) {
          gwas_status <- "allele_mismatch"
        } else {
          if (gwas_beta < 0) {
            protective_allele <- gwas_a1_u
            risk_allele <- gwas_a2_u
            gwas_status <- "a1_protective"
          } else {
            protective_allele <- gwas_a2_u
            risk_allele <- gwas_a1_u
            gwas_status <- "a2_protective"
          }
          if (risk_allele == pvar_alt_u) {
            risk_sign <- 1L
          } else if (risk_allele == pvar_ref_u) {
            risk_sign <- -1L
          } else {
            gwas_status <- "allele_mismatch"
          }
        }
      }
    }

    ## choose dosage and labels from gwas risk-sign mode or fallback variant_id mode
    use_gwas_orientation <- isTRUE(gwas_eff) && !is.na(risk_sign)
    if (use_gwas_orientation) {
      geno_dosage_plot <- if (risk_sign == -1L) 2 - geno_dosage_raw else geno_dosage_raw
      geno_labels <- make_geno_labels(geno_dosage_plot, protective_allele, risk_allele)
      orientation_mode <- "gwas_risk_sign"
    } else {
      geno_dosage_plot <- if (!is.na(ori_sign) && ori_sign == -1L) 2 - geno_dosage_raw else geno_dosage_raw
      geno_labels <- make_geno_labels(geno_dosage_plot, id_ref, id_alt)
      orientation_mode <- "variant_id_refalt"
    }

    # Order genotypes by dosage within this variant
    geno_order <- unique(geno_labels[order(geno_dosage_plot)])
    geno_order <- geno_order[!is.na(geno_order)]
    # Build facet label
    var_label <- if (!is.na(rsid_val) && nzchar(rsid_val)) {
      sprintf("%s (%s)", var_id, rsid_val)
    } else {
      var_id
    }
    facet_label <- sprintf("%s\n%s", var_label, gene_name)
    ## set beta orientation source to match plotted risk dosage axis
    beta_raw <- if ("slope" %in% names(eqtl_pairs)) eqtl_pairs$slope[i] else NA
    beta_val <- if (use_gwas_orientation) orient_effect(beta_raw, risk_sign) else orient_effect(beta_raw, ori_sign)
    pval_val <- if ("pval_nominal" %in% names(eqtl_pairs)) eqtl_pairs$pval_nominal[i] else NA
    data.table(
      sample_id = names(expr_vals),
      expression = as.numeric(expr_vals),
      genotype = factor(geno_labels, levels = geno_order),  # Factor per-variant
      dosage = geno_dosage_plot,
      dosage_raw = geno_dosage_raw,
      DX = sample_info$DX,
      gene_id = gene_id,
      gene_name = gene_name,
      variant_id = var_id,
      facet = facet_label,
      orientation_status = ori_status,
      orientation_sign = ori_sign,
      gwas_A1 = gwas_a1,
      gwas_A2 = gwas_a2,
      gwas_beta = gwas_beta,
      gwas_status = gwas_status,
      risk_sign = risk_sign,
      protective_allele = protective_allele,
      risk_allele = risk_allele,
      orientation_mode = orientation_mode,
      beta = beta_val,
      pval = pval_val
    )
  })
  plot_dt <- rbindlist(plot_list)
  ## warn once when gwas mode falls back to variant_id orientation
  if (isTRUE(gwas_eff)) {
    fallback_status <- c("missing_gwas_cols", "missing_pvar", "missing_gwas", "missing_beta", "zero_beta", "allele_mismatch")
    fallback_dt <- unique(plot_dt[gwas_status %in% fallback_status, .(variant_id, gwas_status)])
    if (nrow(fallback_dt) > 0L) {
      show_n <- min(5L, nrow(fallback_dt))
      show_txt <- paste(sprintf("%s (%s)", fallback_dt$variant_id[seq_len(show_n)], fallback_dt$gwas_status[seq_len(show_n)]), collapse = ", ")
      warning(sprintf(
        "gwas_eff=TRUE fell back to variant_id REF/ALT orientation for %d variant(s) in risk-sign mode: %s%s",
        nrow(fallback_dt),
        show_txt,
        if (nrow(fallback_dt) > show_n) ", ..." else ""
      ))
    }
  }
  ## set x-axis label with effect allele only when gwas risk-sign orientation was used
  x_axis_label <- "Genotype"
  if (isTRUE(gwas_eff)) {
    effect_alleles <- unique(plot_dt[orientation_mode == "gwas_risk_sign" & !is.na(risk_allele) & nzchar(risk_allele), risk_allele])
    if (length(effect_alleles) == 1L) {
      x_axis_label <- sprintf("Genotype (effect allele: %s)", effect_alleles[1])
    } else if (length(effect_alleles) > 1L) {
      x_axis_label <- "Genotype (effect allele: varies by facet)"
    }
  }
  ## resolve auto title from ds context and plotted gene(s)
  resolve_context_label <- function(ds_value) {
    if (is.null(ds_value)) return(NULL)
    ds_chr <- as.character(ds_value)[1]
    if (!nzchar(ds_chr)) return(NULL)
    if (exists("ds2name", inherits = TRUE)) {
      lut <- get("ds2name", inherits = TRUE)
      if (!is.null(names(lut)) && ds_chr %in% names(lut)) return(unname(lut[[ds_chr]]))
    }
    ds_chr
  }
  gene_title <- unique(na.omit(as.character(eqtl_pairs$gene_name)))
  gene_title <- gene_title[nzchar(gene_title)]
  gene_title <- if (length(gene_title) == 1L) gene_title else "multiple genes"
  context_label <- resolve_context_label(ds)
  auto_title <- if (!is.null(context_label)) sprintf("%s %s expression", context_label, gene_title) else sprintf("%s expression", gene_title)
  title_to_use <- NULL
  if (!is.null(title) && length(title) > 0L) {
    title_chr <- as.character(title)[1]
    if (nzchar(title_chr)) title_to_use <- if (identical(title_chr, "auto")) auto_title else title_chr
  }
  # ---- Create annotation labels for each facet ----
  annot_dt <- plot_dt[, .(
    beta = unique(beta),
    pval = unique(pval),
    ymin = min(expression),
    n_geno = length(unique(genotype))
  ), by = facet]
  annot_dt[, label := sprintf("beta = %.3g\np = %.2e", beta, pval)]
  # Position at rightmost genotype (x) and bottom of facet (y)
  annot_dt[, x := n_geno]
  annot_dt[, y := ymin]
  # ---- Build plot ----
  p <- ggplot(plot_dt, aes(x = genotype, y = expression)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5) +
    geom_jitter(aes(color = DX), width = 0.2, size = 2, alpha = 0.8) +
    facet_wrap(~ facet, scales = "free") +
    scale_color_manual(values = c("NTC" = "#1b9e77", "SCZ" = "#d95f02")) +
    scale_x_discrete(drop = TRUE) +
    labs(
      x = x_axis_label,
      y = ifelse(use_raw, "Expression (TMM log2 CPM)", "Residualized Expression"),
      color = "Dx"
      #caption = "Genotype and beta are shown in reference-genome (variant_id) orientation."
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 9),
      legend.position = "bottom"
    )

  # Add beta/p-value annotations
  if (all(!is.na(annot_dt$beta))) {
    p <- p + geom_text(data = annot_dt,
                       aes(x = x+0.8, y = y, label = label),
                       hjust = 1.1, vjust = 0.9, size = 3, inherit.aes = FALSE)
  }
  if (!is.null(title_to_use)) {
    p <- p + ggtitle(title_to_use)
  }
  p
}

```


```{r show MAPK3 in neuropil}
##ds <- 'pnn' ## actually none of the other datasets have it
## apply(indres, \(x) x[gene_name=="MAPK3"]) --> only shows it in neuropil
#dsname <- unname(ds2name[ds])
## load data for the specified dataset
ds <- 'neuropil'
npil_spe <- qs_read(file.path(tqtl_in, sprintf('%s.gene.spe.qs2',ds)))
## use assays(spe)$tmm for expression data, TMM normalized log(CPM+1)
npil_fcovars <- file.path(tqtl_in, sprintf("%s.gene.covars.txt", ds))
npil_plotdata <- prep_eQTLPlotData(npil_spe, geno, npil_fcovars)

indr <- indres[[ds]]
stopifnot(!is.null(indr))

ieqtls <- indr[gene_name=="MAPK3"]
stopifnot(nrow(ieqtls)>0)
#top_eqtls[, .(phenotype_id, variant_id, slope, pval_nominal, pval_perm, af)]

# Create plot
## this might have to be customized further
p <- plot_eQTL_boxplots(npil_plotdata, ieqtls, title=ds2name[ds], gwas_eff = TRUE)
print(p)
```

```{r plot from nominal results in neun dataset}
ds <- 'neun'
neun_spe <- qs_read(file.path(tqtl_in, sprintf('%s.gene.spe.qs2',ds)))
## use assays(spe)$tmm for expression data, TMM normalized log(CPM+1)
neun_fcovars <- file.path(tqtl_in, sprintf("%s.gene.covars.txt", ds))
neun_plotdata <- prep_eQTLPlotData(neun_spe, geno, neun_fcovars)

eqtls <- nomres[["neun"]][gene_name=="MAPK3" & variant_id == "chr16:30114519:G:C"]
pnom <- plot_eQTL_boxplots(neun_plotdata, eqtls, gwas_eff = TRUE)
print(pnom)
```

```{r plot KANSL1-AS1 rs1378358 in Neuropil}
#nvec <- sapply(indres, \(x) nrow(x[gene_name=="KANSL1-AS1"]))
ds="neuropil"
spe <- qs_read(file.path(tqtl_in, sprintf('%s.gene.spe.qs2',ds)))
fcovars <- file.path(tqtl_in, sprintf("%s.gene.covars.txt", ds))
plotdata <- prep_eQTLPlotData(spe, geno, fcovars)
eqtls <- nomres[[ds]][gene_name=="KANSL1-AS1" & rsid== "rs1378358"]

pl <- plot_eQTL_boxplots(plotdata, eqtls)
print(pl)
#plx <- plot_eQTL_boxplots(plotdata, eqtls, restrict_donors = xdonors)
#print(plx)

```

```{r plot KANSL1-AS1 gene}
## which indep results have KANSL1-AS1?
nvec <- sapply(indres, \(x) nrow(x[gene_name=="KANSL1-AS1"]))
ctxs <- names(nvec[nvec>0])
message("Contexts with KANSL1-AS1 independent eQTLs: ", paste(ctxs, collapse=", "))
for (ds in ctxs) {
  dsname <- unname(ds2name[ds])
  spe <- qs_read(file.path(tqtl_in, sprintf('%s.gene.spe.qs2',ds)))
  covar_file <- file.path(tqtl_in, sprintf("%s.gene.covars.txt", ds))
  plotdata <- prep_eQTLPlotData(spe, geno, covar_file)
  indr <- indres[[ds]]
  ieqtls <- indr[gene_name=="KANSL1-AS1"]
  p <- plot_eQTL_boxplots(plotdata, ieqtls, title = sprintf("%s KANSL1-AS1 eQTL", dsname))
  print(p)
}
```


```{r neuropil example}
ds <- 'neuropil'
dsname <- unname(ds2name[ds])
spe <- qs_read(file.path(tqtl_in, sprintf('%s.gene.spe.qs2',ds)))
## use assays(spe)$tmm for expression data, TMM normalized log(CPM+1)
covar_file <- file.path(tqtl_in, sprintf("%s.gene.covars.txt", ds))
indr <- indres[[ds]]
stopifnot(!is.null(indr))
plotdata <- prep_eQTLPlotData(spe, geno, covar_file)
```

```{r}
top_eqtls <- indr[phenotype_id=="ENSG00000102882"]
cat("Selected eQTLs for plotting:\n")
top_eqtls[, .(phenotype_id, variant_id, slope, pval_nominal, pval_perm, af)]
print(plot_eQTL_boxplots(plotdata, top_eqtls))
```
