---
title: "final coloc overlap exploration"
output: html_notebook
---

```{r setup}
suppressMessages({
  library(data.table)
  library(ggplot2)
  library(stringr)
  library(openxlsx)
  library(VennDiagram)
  library(grid)
  library(here)
})

##  set project paths and key inputs for dlpfc and habenula coloc summaries.
here::i_am(".git/HEAD")
source(here("code", "analysis", "15_eqtl_coloc", "table_utils.R"))
datadir <- here("processed-data", "eQTL")
refdir <- here("processed-data", "ref")
hb_root <- "/home/gpertea/work/R/Habenula_Pilot"

f_dlpfc <- file.path(datadir, "coloc", "coloc_abf_results.tsv.gz")
f_dlpfc_gated <- file.path(datadir, "coloc", "coloc_abf_results_gated.tsv.gz")
f_hb_source <- file.path(hb_root, "processed-data", "18_coloc", "supp_table.csv")
hb_gene_dir <- here("code", "analysis", "15_eqtl_coloc", "gene_lists")
f_hb_gene_list <- file.path(hb_gene_dir, "habenulaPilot-coloc-genes.csv")
f_ext_meta <- file.path(hb_gene_dir, "external_reference_metadata.tsv")
f_ref85_scz <- file.path(hb_gene_dir, "ref85-scz_only_derived.csv")
f_ref86 <- file.path(hb_gene_dir, "ref86-devbrain-coloc.csv")
f_ref87_int18 <- file.path(hb_gene_dir, "ref87-INT18-PsychEncode-SCZ-risk.csv")
f_dlpfc_donor <- file.path(refdir, "donor_meta.tsv.gz")
f_hb_coldata <- file.path(hb_root, "processed-data", "17_eQTL", "tensorQTL_input", "colData.csv")
f_sens_sum <- file.path(datadir, "coloc", "prior_sensitivity", "w500k_prior_sensitivity_summary.tsv.gz")
f_sens_gene <- file.path(datadir, "coloc", "prior_sensitivity", "w500k_prior_sensitivity_per_gene.tsv.gz")
f_sens_fail <- file.path(datadir, "coloc", "prior_sensitivity", "w500k_prior_sensitivity_fail_summary.tsv.gz")
f_sens_fail_table <- file.path(datadir, "coloc", "prior_sensitivity", "w500k_prior_sensitivity_fail_table.tsv.gz")

##  overlap table specs for reproducible slide-ready exports.
overlap_specs <- list(
  gene_variant = c("gene_symbol_key", "rsid_key"),
  gene_only = c("gene_symbol_key")
)
stopifnot(identical(sort(names(overlap_specs)), c("gene_only", "gene_variant")))

##  define output location for final comparison artifacts.
final_dir <- file.path(datadir, "coloc", "final")
if (!dir.exists(final_dir)) dir.create(final_dir, recursive = TRUE)
```

```{r concept-notes, eval=FALSE}
##  prior-sensitivity in coloc.abf means checking how stable posterior conclusions are when priors vary, especially p12.
##  this is recommended because posterior probabilities can shift under different prior beliefs about shared causality.
##  our base coloc run for dlpfc used fixed priors p1=1e-4, p2=1e-4, p12=1e-5 and then sensitivity varied p12 over a grid.
##  habenula used coloc.abf without explicit prior arguments, so coloc defaults were used (the same default values above).
##  dlpfc strategy differs by adding dual-window evaluation (500kb primary plus 1mb sensitivity) and a separate prior-sensitivity report.
##  habenula strategy centered on the primary coloc run and downstream strong filtering at pp.h4.abf > 0.8.
```

```{r helpers}
##  define small helper utilities for schema checks and robust aggregation.
partition_levels <- c(
  "spd_spg_hb", "spd_spg_only", "spd_hb_only", "spg_hb_only",
  "spd_only", "spg_only", "hb_only"
)

std_rsid_key <- function(x) {
  y <- tolower(trimws(as.character(x)))
  y[y == "" | is.na(y)] <- NA_character_
  y
}

add_overlap_key <- function(dt, key_fields, key_sep = "||") {
  d <- copy(dt)
  for (k in key_fields) {
    d <- d[!is.na(get(k)) & get(k) != ""]
  }
  d[, overlap_key := do.call(paste, c(.SD, sep = key_sep)), .SDcols = key_fields]
  d
}

collapse_set_rows <- function(dt, overlap_type, key_fields, set_name) {
  d <- add_overlap_key(dt, key_fields = key_fields)
  d[, .(
    overlap_type = overlap_type,
    gene_name = first_non_missing(gene_name),
    rsid = first_non_missing(rsid),
    source_set = set_name,
    contexts = uniq_concat(context_or_group),
    gene_name_list = uniq_concat(gene_name),
    rsid_list = uniq_concat(rsid)
  ), by = overlap_key]
}

collapse_annotation_rows <- function(dt, overlap_type, key_fields) {
  d <- add_overlap_key(dt, key_fields = key_fields)
  d[, .(
    overlap_type = overlap_type,
    gene_name = first_non_missing(gene_name),
    rsid = first_non_missing(rsid),
    source_set = uniq_concat(source_set),
    contexts = uniq_concat(context_or_group),
    gene_name_list = uniq_concat(gene_name),
    rsid_list = uniq_concat(rsid)
  ), by = overlap_key]
}

complete_partition_counts <- function(partitions_dt, overlap_type) {
  obs <- partitions_dt[, .N, by = partition_label]
  tmpl <- data.table(partition_label = partition_levels)
  out <- merge(tmpl, obs, by = "partition_label", all.x = TRUE)
  out[is.na(N), N := 0L]
  out[, overlap_type := overlap_type]
  setcolorder(out, c("overlap_type", "partition_label", "N"))
  out
}

partition_to_venn_counts <- function(part_dt) {
  x <- setNames(part_dt$N, part_dt$partition_label)
  get_n <- function(k) if (k %chin% names(x)) as.integer(x[[k]]) else 0L
  spd_spg_hb <- get_n("spd_spg_hb")
  spd_spg_only <- get_n("spd_spg_only")
  spd_hb_only <- get_n("spd_hb_only")
  spg_hb_only <- get_n("spg_hb_only")
  spd_only <- get_n("spd_only")
  spg_only <- get_n("spg_only")
  hb_only <- get_n("hb_only")
  list(
    area1 = spd_only + spd_spg_only + spd_hb_only + spd_spg_hb,
    area2 = spg_only + spd_spg_only + spg_hb_only + spd_spg_hb,
    area3 = hb_only + spd_hb_only + spg_hb_only + spd_spg_hb,
    n12 = spd_spg_only + spd_spg_hb,
    n13 = spd_hb_only + spd_spg_hb,
    n23 = spg_hb_only + spd_spg_hb,
    n123 = spd_spg_hb
  )
}

draw_venn_counts <- function(venn_counts, title_label, out_png = NULL) {
  if (!is.null(out_png)) {
    png(out_png, width = 1400, height = 1000, res = 140)
    on.exit(dev.off(), add = TRUE)
  }
  grid.newpage()
  vp <- VennDiagram::draw.triple.venn(
    area1 = venn_counts$area1,
    area2 = venn_counts$area2,
    area3 = venn_counts$area3,
    n12 = venn_counts$n12,
    n13 = venn_counts$n13,
    n23 = venn_counts$n23,
    n123 = venn_counts$n123,
    category = c("7 SpDs", "4 SPGs", "habenulaPilot"),
    fill = c("#8cc0de", "#6e9dc0", "#c2d17d"),
    cex = 1.1, cat.cex = 1.0, cat.pos = c(-20, 20, 180),
    euler.d = FALSE, scaled = FALSE, ind = FALSE
  )
  grid.draw(vp)
  grid.text(
    label = title_label,
    x = unit(0.5, "npc"), y = unit(0.98, "npc"),
    gp = gpar(fontsize = 11, fontface = "bold")
  )
}
```

```{r load-inputs}
##  load required dlpfc and habenula coloc tables and assert expected schema.
assert_file(f_dlpfc)
assert_file(f_dlpfc_gated)
assert_file(f_hb_source)
assert_file(f_dlpfc_donor)
assert_file(f_hb_coldata)

dlpfc <- fread(f_dlpfc)
dlpfc_gated <- fread(f_dlpfc_gated)
hb <- fread(f_hb_source)

##  cache compact habenula coloc table locally (gene_name + variant_id) when missing.
if (!dir.exists(hb_gene_dir)) dir.create(hb_gene_dir, recursive = TRUE)
hb_gene_list <- tryCatch(fread(f_hb_gene_list), error = function(e) NULL)
if (
  is.null(hb_gene_list) ||
  !all(c("gene_name", "variant_id", "rsID") %chin% names(hb_gene_list))
) {
  hb_gene_list <- unique(hb[, .(
    gene_name = as.character(gene_symbol),
    variant_id = as.character(snp),
    rsID = as.character(snp_rs_id)
  )])
  hb_gene_list <- hb_gene_list[
    !is.na(gene_name) & gene_name != "" &
    !is.na(variant_id) & variant_id != ""
  ]
  hb_gene_list[is.na(rsID) | rsID == "", rsID := ""]
  setorder(hb_gene_list, gene_name, variant_id, rsID)
  hb_gene_list <- unique(hb_gene_list, by = c("gene_name", "variant_id", "rsID"))
  fwrite(hb_gene_list, f_hb_gene_list, sep = ",", quote = TRUE)
}
hb_gene_list <- fread(f_hb_gene_list)
assert_cols(hb_gene_list, c("gene_name", "variant_id", "rsID"), "habenulaPilot-coloc-genes.csv")

assert_cols(dlpfc, c("fid", "gene_name", "lead_snp", "rsid", "PP4", "ds"), "dlpfc coloc table")
assert_cols(dlpfc_gated, c("window", "ds", "fid", "gene_name", "cat_gated", "gate_pass", "gate_reason"), "dlpfc gated coloc table")
assert_cols(hb, c("gene", "gene_symbol", "snp", "snp_rs_id", "PP.H4.abf"), "habenula supp table")
```

```{r gated-pass-clean-export}
##  build pass-only cleaned coloc results and add binary gwas/dge/eqtl annotations for reporting.
f_eqtl_cis <- file.path(datadir, "tqtl_out", "cis_independent_FDR05_merged.xlsx")
f_eqtl_nom <- file.path(datadir, "tqtl_out", "nominal_BH05_merged.xlsx")
f_gwas_match <- file.path(refdir, "GWAS_SCZ_tqtl-matched.tab.gz")
assert_file(f_eqtl_cis)
assert_file(f_eqtl_nom)
assert_file(f_gwas_match)

gated_pass <- dlpfc_gated[gate_reason == "pass"]
assert_cols(
  gated_pass,
  c("ds", "fid", "gene_name", "lead_snp", "rsid", "nsnps", "PP3", "PP4", "PP34", "PP4_over_PP34", "cat_gated"),
  "gated pass coloc table"
)
gated_pass[, cat := as.character(cat_gated)]

##  gwas risk flag from harmonized gwas variants at p < 5e-8.
gwas_match <- fread(f_gwas_match, select = c("variant_id", "p"))
gwas_risk_set <- unique(gwas_match[p < 5e-8 & !is.na(variant_id), variant_id])
gated_pass <- apply_gwas_flag(gated_pass, variant_col = "lead_snp", gwas_variants = gwas_risk_set, out_col = "GWAS")

##  eqtl/eqtlnom flags by context + variant using merged canonical workbooks.
cis_merged <- as.data.table(read.xlsx(f_eqtl_cis, sheet = 1, detectDates = FALSE))
nom_merged <- as.data.table(read.xlsx(f_eqtl_nom, sheet = 1, detectDates = FALSE))
assert_cols(cis_merged, c("expression_context", "variant_id"), "cis merged workbook")
assert_cols(nom_merged, c("expression_context", "variant_id"), "nominal merged workbook")
cis_key <- unique(cis_merged[!is.na(expression_context) & !is.na(variant_id), .(
  ds = as.character(expression_context),
  lead_snp = as.character(variant_id)
)])
nom_key <- unique(nom_merged[!is.na(expression_context) & !is.na(variant_id), .(
  ds = as.character(expression_context),
  lead_snp = as.character(variant_id)
)])
gated_pass[, eQTL := 0L]
gated_pass[, eQTLnom := 0L]
gated_pass[cis_key, on = .(ds, lead_snp), eQTL := 1L]
gated_pass[nom_key, on = .(ds, lead_snp), eQTLnom := 1L]

##  dge flag by context + gene using layer-adjusted spd and context-specific spg files.
deg_sets <- load_deg_sets(contexts = default_eqtl_contexts, project_root = here(), rdsdir = here("processed-data", "rds"))
gated_pass <- apply_deg_flags(gated_pass, context_col = "ds", gene_col = "fid", deg_sets = deg_sets)

coloc_clean_pass <- gated_pass[, .(
  ds, fid, gene_name, lead_snp, rsid, nsnps,
  PP3, PP4, PP34, PP4_over_PP34, cat,
  GWAS, DGE, DGEp05, eQTL, eQTLnom
)]
setorder(coloc_clean_pass, ds, cat, -PP4, -PP34, fid)
validate_binary_cols(coloc_clean_pass, c("GWAS", "DGE", "DGEp05", "eQTL", "eQTLnom"), "coloc_clean_pass")
if (coloc_clean_pass[DGE == 1L & DGEp05 != 1L, .N] > 0L) stop("coloc_clean_pass has DGE==1 rows not present in DGEp05.")

f_coloc_clean_tsv <- file.path(final_dir, "coloc_abf_results_gated_pass_clean.tsv.gz")
f_coloc_clean_xlsx <- file.path(datadir, "coloc", "coloc_results.xlsx")
fwrite(coloc_clean_pass, f_coloc_clean_tsv, sep = "\t", quote = FALSE)

coloc_clean_pass_xlsx <- rename_coloc_excel_cols(coloc_clean_pass)
setcolorder(coloc_clean_pass_xlsx, c(
  "expression_context", "ENSEMBL", "gene_name", "lead_snp", "rsid", "nsnps",
  "PP3", "PP4", "PP34", "PP4_over_PP34", "hypothesis_category",
  "risk_SNP_GWAS", "SCZD_DGE_layer_adjusted", "SCZD_DGE_layer_adjusted_p05",
  "eQTL_independent", "eQTL_nominal"
))

wb <- createWorkbook()
addWorksheet(wb, "coloc_pass")
writeData(wb, "coloc_pass", coloc_clean_pass_xlsx)
saveWorkbook(wb, f_coloc_clean_xlsx, overwrite = TRUE)

coloc_clean_pass
```

```{r canonical-schema}
##  build canonical per-row schema so all downstream outputs have consistent annotation fields.
dlpfc_can <- dlpfc[, .(
  source_project = "DLPFC",
  source_set = fifelse(grepl("^spd", ds), "spd_union_input", "spg_union_input"),
  context_or_group = as.character(ds),
  group = fifelse(grepl("^spd", ds), "SpD", "SPG"),
  gene_id_raw = as.character(fid),
  gene_id_norm = normalize_gene_id(fid),
  gene_name = as.character(gene_name),
  variant_id = as.character(lead_snp),
  rsid = as.character(rsid),
  PP4 = as.numeric(PP4),
  cat = if ("cat" %chin% names(dlpfc)) as.character(cat) else NA_character_
)]
dlpfc_can[rsid == "", rsid := NA_character_]
dlpfc_can[, gene_symbol_key := std_symbol_key(gene_name)]
dlpfc_can[, rsid_key := std_rsid_key(rsid)]

hb_can <- hb[, .(
  source_project = "habenulaPilot",
  source_set = "habenula_input",
  context_or_group = "habenulaPilot",
  group = "habenulaPilot",
  gene_id_raw = as.character(gene),
  gene_id_norm = normalize_gene_id(gene),
  gene_name = as.character(gene_symbol),
  variant_id = as.character(snp),
  rsid = as.character(snp_rs_id),
  PP4 = as.numeric(`PP.H4.abf`)
)]
hb_can[rsid == "", rsid := NA_character_]
hb_can[, gene_symbol_key := std_symbol_key(gene_name)]
hb_can[, rsid_key := std_rsid_key(rsid)]
```

```{r strong-filter}
##  define strong coloc sets using pp4 > 0.8 for cross-study comparability.
strong_dlpfc <- dlpfc_can[PP4 > 0.8]
strong_hb <- hb_can[PP4 > 0.8]

##  check optional consistency between pp4 thresholding and existing cat labels in dlpfc.
if ("cat" %chin% names(strong_dlpfc)) {
  dlpfc_cat_check <- dlpfc_can[!is.na(cat), .(
    n = .N,
    n_pp4_strong = sum(PP4 > 0.8),
    n_cat_strong = sum(cat == "strong_coloc")
  )]
  dlpfc_cat_check
}

##  bind all strong rows to simplify multi-set annotation and symbol ambiguity diagnostics.
all_strong <- rbindlist(list(strong_dlpfc, strong_hb), fill = TRUE, use.names = TRUE)
```

```{r overlap-build-and-export}
##  construct spd, spg, and habenula overlap tables using two explicit overlap types.
overlap_results <- list()
partition_counts_all <- list()

spd_rows <- strong_dlpfc[group == "SpD"]
spg_rows <- strong_dlpfc[group == "SPG"]
hb_rows <- strong_hb

for (overlap_type in names(overlap_specs)) {
  key_fields <- overlap_specs[[overlap_type]]

  spd_set <- collapse_set_rows(spd_rows, overlap_type = overlap_type, key_fields = key_fields, set_name = "spd_union_input")
  spg_set <- collapse_set_rows(spg_rows, overlap_type = overlap_type, key_fields = key_fields, set_name = "spg_union_input")
  hb_set <- collapse_set_rows(hb_rows, overlap_type = overlap_type, key_fields = key_fields, set_name = "habenula_input")

  all_keys <- sort(unique(c(spd_set$overlap_key, spg_set$overlap_key, hb_set$overlap_key)))
  venn_core <- data.table(overlap_type = overlap_type, overlap_key = all_keys)
  venn_core[, in_spd := overlap_key %chin% spd_set$overlap_key]
  venn_core[, in_spg := overlap_key %chin% spg_set$overlap_key]
  venn_core[, in_hb := overlap_key %chin% hb_set$overlap_key]
  venn_core[, partition_label := fcase(
    in_spd & in_spg & in_hb, "spd_spg_hb",
    in_spd & in_spg & !in_hb, "spd_spg_only",
    in_spd & !in_spg & in_hb, "spd_hb_only",
    !in_spd & in_spg & in_hb, "spg_hb_only",
    in_spd & !in_spg & !in_hb, "spd_only",
    !in_spd & in_spg & !in_hb, "spg_only",
    !in_spd & !in_spg & in_hb, "hb_only",
    default = "none"
  )]

  ann <- collapse_annotation_rows(all_strong, overlap_type = overlap_type, key_fields = key_fields)
  overlap_table <- merge(venn_core, ann, by = c("overlap_type", "overlap_key"), all.x = TRUE)
  overlap_table[, partition_label := factor(partition_label, levels = partition_levels)]
  setorder(overlap_table, partition_label, overlap_key)

  overlap_table[, partition_label := as.character(partition_label)]
  overlap_table[, contexts := fifelse(is.na(contexts), "", contexts)]
  overlap_table[, gene_name := fifelse(is.na(gene_name), "", gene_name)]
  overlap_table[, rsid := fifelse(is.na(rsid), "", rsid)]
  overlap_table[, rsid_list := fifelse(is.na(rsid_list), "", rsid_list)]

  setcolorder(overlap_table, c(
    "overlap_type", "overlap_key", "partition_label",
    "in_spd", "in_spg", "in_hb",
    "gene_name", "rsid", "rsid_list", "contexts",
    "source_set", "gene_name_list"
  ))

  out_name <- if (overlap_type == "gene_variant") {
    "strong_coloc_overlap_table_gene_variant.tsv.gz"
  } else {
    "strong_coloc_overlap_table_gene_only.tsv.gz"
  }
  fwrite(overlap_table, file.path(final_dir, out_name), sep = "\t", quote = FALSE)

  group_counts <- data.table(
    overlap_type = overlap_type,
    source_set = c("spd_union_input", "spg_union_input", "habenula_input"),
    n_keys = c(uniqueN(spd_set$overlap_key), uniqueN(spg_set$overlap_key), uniqueN(hb_set$overlap_key))
  )

  pcounts <- complete_partition_counts(overlap_table, overlap_type = overlap_type)
  partition_counts_all[[overlap_type]] <- pcounts

  overlap_results[[overlap_type]] <- list(
    overlap_table = overlap_table,
    group_counts = group_counts,
    partition_counts = pcounts
  )
}

partition_counts_all <- rbindlist(partition_counts_all, fill = TRUE, use.names = TRUE)
f_part_counts <- file.path(final_dir, "strong_coloc_partition_counts_by_overlap_type.tsv.gz")
fwrite(partition_counts_all, f_part_counts, sep = "\t", quote = FALSE)
```

```{r validation-checks}
##  run basic validation checks on overlap outputs and required metadata presence.
for (overlap_type in names(overlap_specs)) {
  x <- overlap_results[[overlap_type]]$overlap_table
  stopifnot(nrow(x) > 0L)
  stopifnot(any(x$in_spd | x$in_spg | x$in_hb))
  stopifnot(uniqueN(x$overlap_key) == nrow(x))
}

##  enforce gene-variant key completeness and uniqueness.
gv <- overlap_results[["gene_variant"]]$overlap_table
stopifnot(gv[gene_name == "" | is.na(gene_name), .N] == 0L)
stopifnot(gv[rsid == "" | is.na(rsid), .N] == 0L)
stopifnot(uniqueN(gv$overlap_key) == nrow(gv))

##  expected gene-only partition counts for this analysis snapshot.
expected_gene_only <- data.table(
  partition_label = partition_levels,
  expected_N = c(1L, 4L, 0L, 0L, 10L, 2L, 15L)
)
observed_gene_only <- partition_counts_all[overlap_type == "gene_only", .(partition_label, N)]
chk <- merge(expected_gene_only, observed_gene_only, by = "partition_label", all.x = TRUE)
chk[is.na(N), N := 0L]
stopifnot(all(chk$expected_N == chk$N))
chk
```

```{r venn-plots}
##  save and show one stable 3-way venn for the gene-only overlap.
gene_only_part <- partition_counts_all[overlap_type == "gene_only", .(partition_label, N)]
gene_only_venn_counts <- partition_to_venn_counts(gene_only_part)
f_gene_only_venn <- file.path(final_dir, "strong_coloc_venn_gene_only.png")
draw_venn_counts(
  venn_counts = gene_only_venn_counts,
  title_label = "strong coloc overlap (gene_only mode; PP4 > 0.8)",
  out_png = f_gene_only_venn
)
draw_venn_counts(
  venn_counts = gene_only_venn_counts,
  title_label = ""
)
```

```{r context-barplot}
##  summarize gated strong coloc genes by context and export a plot-ready table.
ctx_map <- c(
  spd01 = "SpD01-WMtz",
  spd02 = "SpD02-L3/4",
  spd03 = "SpD03-L6",
  spd04 = "SpD04-WM",
  spd05 = "SpD05-L5",
  spd06 = "SpD06-L2/3",
  spd07 = "SpD07-L1-M",
  neun = "Neun",
  neuropil = "Neuropil",
  pnn = "PNN",
  vasc = "Vasc"
)
ctx_levels <- names(ctx_map)
ctx_labels <- unname(ctx_map)

strong_gated <- dlpfc_gated[cat_gated == "strong_coloc", .(ds, fid, gene_name)]
ctx_summary <- strong_gated[, .(
  n_strong = uniqueN(fid),
  gene_list = paste(sort(unique(gene_name[!is.na(gene_name) & gene_name != ""])), collapse = ", ")
), by = ds]

ctx_summary <- data.table(ds = ctx_levels)[ctx_summary, on = "ds"]
ctx_summary[is.na(n_strong), n_strong := 0L]
ctx_summary[is.na(gene_list), gene_list := ""]
ctx_summary[, context_label := unname(ctx_map[ds])]
ctx_summary[is.na(context_label), context_label := ds]
ctx_summary[, context_label := factor(context_label, levels = rev(ctx_labels))]
ctx_summary[, gene_list_wrapped := fifelse(gene_list == "", "", str_wrap(gene_list, width = 44))]

f_ctx_strong <- file.path(final_dir, "strong_coloc_gated_by_context.tsv.gz")
fwrite(
  ctx_summary[, .(ds, context_label = as.character(context_label), n_strong, gene_list)],
  f_ctx_strong,
  sep = "\t",
  quote = FALSE
)

ggplot(ctx_summary, aes(x = context_label, y = n_strong)) +
  geom_col(width = 0.4, fill = "#4f83b6") +
  geom_text(
    aes(y = n_strong + 0.08, label = ifelse(n_strong > 0, as.character(n_strong), "")),
    hjust = 0,
    size = 4.5
  ) +
  #geom_text( aes(y = n_strong + 0.55, label = gene_list_wrapped),
  #  hjust = 0, size = 2.9,
  #  lineheight = 0.95 ) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.62))) +
  theme_bw(base_size = 14) +
  labs(
    x = "context",
    y = "number of gene-variant pairs",
    title = "Strong colocalization by context (PP4>0.8) "
    #subtitle = "PP4 > 0.8"
  ) +
  theme(
    axis.text.y = element_text(size = 14),
    plot.margin = margin(6, 160, 6, 6, unit = "pt")
  )
```

```{r group-barplot}
##  compare spd/spg/habenula set sizes for each overlap type.
group_counts_all <- rbindlist(lapply(names(overlap_specs), function(k) overlap_results[[k]]$group_counts), fill = TRUE)
group_counts_all[, source_set := factor(source_set, levels = c("spd_union_input", "spg_union_input", "habenula_input"))]

ggplot(group_counts_all, aes(x = source_set, y = n_keys, fill = source_set)) +
  geom_col(width = 0.75) +
  facet_wrap(~ overlap_type, ncol = 1) +
  theme_bw() +
  labs(
    x = "grouped set",
    y = "unique strong overlap keys",
    title = "strong set sizes by overlap type",
    subtitle = "strong definition is PP4 > 0.8"
  ) +
  theme(legend.position = "none")
```

```{r overlap-type-comparison}
##  summarize and plot n_overlap between gene_variant and gene_only by partition class.
type_cmp <- dcast(
  partition_counts_all,
  partition_label ~ overlap_type,
  value.var = "N",
  fill = 0
)
f_type_cmp <- file.path(final_dir, "strong_coloc_n_overlap_type_comparison.tsv.gz")
fwrite(type_cmp, f_type_cmp, sep = "\t")
type_cmp

type_cmp_long <- melt(type_cmp, id.vars = "partition_label", variable.name = "overlap_type", value.name = "N")
ggplot(type_cmp_long, aes(x = partition_label, y = N, fill = overlap_type)) +
  geom_col(position = "dodge") +
  theme_bw() +
  labs(
    x = "intersection class",
    y = "n_overlap",
    title = "n_overlap comparison by overlap type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r donor-overlap}
##  compute donor overlap counts and overlapped ids between dlpfc and habenula cohorts.
dlpfc_donor <- fread(f_dlpfc_donor)
hb_coldata <- fread(f_hb_coldata)
assert_cols(dlpfc_donor, c("BrNum", "DX"), "dlpfc donor metadata")
assert_cols(hb_coldata, c("BrNum"), "habenula colData")

norm_br <- function(x) toupper(trimws(as.character(x)))
d_ids <- sort(unique(norm_br(dlpfc_donor$BrNum)))
h_ids <- sort(unique(norm_br(hb_coldata$BrNum)))
ov_ids <- sort(intersect(d_ids, h_ids))

dx_map <- unique(dlpfc_donor[, .(
  BrNum_norm = norm_br(BrNum),
  Dx = toupper(trimws(as.character(DX)))
)])
donor_overlap_ids <- merge(
  data.table(BrNum_norm = ov_ids),
  dx_map,
  by = "BrNum_norm",
  all.x = TRUE
)
donor_overlap_ids[, BrNum := paste0("Br", sub("^BR", "", BrNum_norm))]
setcolorder(donor_overlap_ids, c("BrNum", "Dx"))
setorder(donor_overlap_ids, Dx, BrNum)
donor_overlap_ids <- donor_overlap_ids[, .(BrNum, Dx)]

fwrite(donor_overlap_ids, file.path(final_dir, "donor_overlap_ids.tsv.gz"), sep = "\t")

donor_overlap_ids
```

```{r cross-study-overlap-inputs}
##  load external reference sets and build the current-study gene set for cross-study overlap.
assert_file(f_hb_gene_list)
assert_file(f_ext_meta)
assert_file(f_ref85_scz)
assert_file(f_ref86)
assert_file(f_ref87_int18)

hb_gl <- fread(f_hb_gene_list)
ext_meta <- fread(f_ext_meta)
ref85 <- fread(f_ref85_scz)
ref86 <- fread(f_ref86, blank.lines.skip = TRUE)
ref87_int18 <- fread(f_ref87_int18)

assert_cols(hb_gl, c("gene_name", "variant_id", "rsID"), "habenulaPilot-coloc-genes.csv")
assert_cols(ext_meta, c(
  "set_name", "paper_short", "source_ref", "evidence_type", "disorder_scope",
  "developmental_stage", "brain_region_group", "brain_regions_detail",
  "n_samples", "n_donors", "ancestry_summary", "demographics_note",
  "comparison_role", "source_url"
), "external_reference_metadata.tsv")
assert_cols(ref85, c("gene_name", "variant", "ref85_selection_rule", "ref85_source_note"), "ref85-scz_only_derived.csv")
assert_cols(ref86, c("gene_name"), "ref86-devbrain-coloc.csv")
assert_cols(ref87_int18, c("sczgenenames"), "ref87-INT18-PsychEncode-SCZ-risk.csv")

to_gene_set <- function(x) {
  y <- std_symbol_key(x)
  y <- y[!is.na(y) & y != ""]
  sort(unique(y))
}

sets5 <- list(
  SCZD_PNN = to_gene_set(strong_dlpfc$gene_name),
  habenulaPilot = to_gene_set(hb_gl$gene_name),
  Zeng2022_ref85_SCZ = to_gene_set(ref85$gene_name),
  Wen2024_ref86 = to_gene_set(ref86$gene_name),
  Wang2018_ref87_INT18 = to_gene_set(ref87_int18$sczgenenames)
)

set_sizes_5way <- data.table(
  set_name = names(sets5),
  n_genes = as.integer(lengths(sets5))
)
stopifnot(all(set_sizes_5way$n_genes > 0L))

message(sprintf(
  "5-way set sizes: %s",
  paste(sprintf("%s=%d", set_sizes_5way$set_name, set_sizes_5way$n_genes), collapse = "; ")
))
set_sizes_5way
```

```{r cross-study-reference-metadata}
##  primary framing: document the external sets before interpreting sparse cross-study overlaps.
study_meta <- ext_meta[
  set_name %chin% c("SCZD_PNN", "habenulaPilot", "Zeng2022_ref85_SCZ", "Wen2024_ref86", "Wang2018_ref87_INT18"),
  .(
    set_name, paper_short, evidence_type, disorder_scope, developmental_stage,
    brain_region_group, brain_regions_detail, n_samples, n_donors,
    ancestry_summary, comparison_role
  )
]
study_meta
```

These comparison sets are intentionally heterogeneous. `habenulaPilot` is the closest prior-study comparator, `Zeng2022_ref85_SCZ` is an adult multi-region SCZ-focused colocalization subset derived from the BREMA schizophrenia rows, `Wen2024_ref86` is a fetal cortical xQTL/colocalization reference, and `Wang2018_ref87_INT18` is a broad PsychENCODE SCZ risk-prioritization framework rather than a coloc-only list.

```{r cross-study-overlap-helpers}
##  helper utilities for membership summaries and venn input counts.
build_membership_table <- function(sets_named) {
  set_names <- names(sets_named)
  universe <- sort(unique(unlist(sets_named, use.names = FALSE)))
  out <- data.table(gene_name = universe)
  for (nm in set_names) {
    out[, (nm) := gene_name %chin% sets_named[[nm]]]
  }
  out[, combo_label := apply(.SD, 1, function(z) {
    present <- set_names[as.logical(z)]
    if (length(present) == 0L) "none" else paste(present, collapse = " & ")
  }), .SDcols = set_names]
  out
}

build_combo_template <- function(set_names) {
  bits <- as.data.table(expand.grid(rep(list(c(FALSE, TRUE)), length(set_names))))
  setnames(bits, set_names)
  bits[, combo_label := apply(.SD, 1, function(z) {
    present <- set_names[as.logical(z)]
    if (length(present) == 0L) "none" else paste(present, collapse = " & ")
  }), .SDcols = set_names]
  bits <- bits[combo_label != "none"]
  bits[, n_sets := rowSums(.SD), .SDcols = set_names]
  setorder(bits, n_sets, combo_label)
  bits
}

summarize_combos <- function(membership_dt, set_names) {
  tmpl <- build_combo_template(set_names)
  obs <- membership_dt[combo_label != "none", .(
    n_genes = .N,
    gene_list = paste(sort(unique(gene_name)), collapse = "; ")
  ), by = combo_label]
  out <- merge(tmpl, obs, by = "combo_label", all.x = TRUE)
  out[is.na(n_genes), n_genes := 0L]
  out[is.na(gene_list), gene_list := ""]
  setcolorder(out, c("combo_label", "n_sets", set_names, "n_genes", "gene_list"))
  setorder(out, -n_genes, -n_sets, combo_label)
  out
}

venn_counts_5way <- function(sets_named) {
  stopifnot(length(sets_named) == 5L)
  s1 <- sets_named[[1]]; s2 <- sets_named[[2]]; s3 <- sets_named[[3]]; s4 <- sets_named[[4]]; s5 <- sets_named[[5]]
  inter_n <- function(...) length(Reduce(intersect, list(...)))
  list(
    area1 = length(s1), area2 = length(s2), area3 = length(s3), area4 = length(s4), area5 = length(s5),
    n12 = inter_n(s1, s2), n13 = inter_n(s1, s3), n14 = inter_n(s1, s4), n15 = inter_n(s1, s5),
    n23 = inter_n(s2, s3), n24 = inter_n(s2, s4), n25 = inter_n(s2, s5),
    n34 = inter_n(s3, s4), n35 = inter_n(s3, s5), n45 = inter_n(s4, s5),
    n123 = inter_n(s1, s2, s3), n124 = inter_n(s1, s2, s4), n125 = inter_n(s1, s2, s5),
    n134 = inter_n(s1, s3, s4), n135 = inter_n(s1, s3, s5), n145 = inter_n(s1, s4, s5),
    n234 = inter_n(s2, s3, s4), n235 = inter_n(s2, s3, s5), n245 = inter_n(s2, s4, s5), n345 = inter_n(s3, s4, s5),
    n1234 = inter_n(s1, s2, s3, s4), n1235 = inter_n(s1, s2, s3, s5),
    n1245 = inter_n(s1, s2, s4, s5), n1345 = inter_n(s1, s3, s4, s5), n2345 = inter_n(s2, s3, s4, s5),
    n12345 = inter_n(s1, s2, s3, s4, s5)
  )
}

venn_counts_4way <- function(sets_named) {
  stopifnot(length(sets_named) == 4L)
  s1 <- sets_named[[1]]; s2 <- sets_named[[2]]; s3 <- sets_named[[3]]; s4 <- sets_named[[4]]
  inter_n <- function(...) length(Reduce(intersect, list(...)))
  list(
    area1 = length(s1), area2 = length(s2), area3 = length(s3), area4 = length(s4),
    n12 = inter_n(s1, s2), n13 = inter_n(s1, s3), n14 = inter_n(s1, s4),
    n23 = inter_n(s2, s3), n24 = inter_n(s2, s4), n34 = inter_n(s3, s4),
    n123 = inter_n(s1, s2, s3), n124 = inter_n(s1, s2, s4), n134 = inter_n(s1, s3, s4), n234 = inter_n(s2, s3, s4),
    n1234 = inter_n(s1, s2, s3, s4)
  )
}

validate_venn_counts_5way <- function(vc) {
  stopifnot(all(unlist(vc) == as.integer(unlist(vc))))
  stopifnot(vc$n12345 <= min(vc$n123, vc$n124, vc$n125, vc$n134, vc$n135, vc$n145, vc$n234, vc$n235, vc$n245, vc$n345))
  stopifnot(vc$n12 <= min(vc$area1, vc$area2))
  stopifnot(vc$n13 <= min(vc$area1, vc$area3))
  stopifnot(vc$n14 <= min(vc$area1, vc$area4))
  stopifnot(vc$n15 <= min(vc$area1, vc$area5))
  stopifnot(vc$n23 <= min(vc$area2, vc$area3))
  stopifnot(vc$n24 <= min(vc$area2, vc$area4))
  stopifnot(vc$n25 <= min(vc$area2, vc$area5))
  stopifnot(vc$n34 <= min(vc$area3, vc$area4))
  stopifnot(vc$n35 <= min(vc$area3, vc$area5))
  stopifnot(vc$n45 <= min(vc$area4, vc$area5))
}

validate_venn_counts_4way <- function(vc) {
  stopifnot(all(unlist(vc) == as.integer(unlist(vc))))
  stopifnot(vc$n1234 <= min(vc$n123, vc$n124, vc$n134, vc$n234))
  stopifnot(vc$n12 <= min(vc$area1, vc$area2))
  stopifnot(vc$n13 <= min(vc$area1, vc$area3))
  stopifnot(vc$n14 <= min(vc$area1, vc$area4))
  stopifnot(vc$n23 <= min(vc$area2, vc$area3))
  stopifnot(vc$n24 <= min(vc$area2, vc$area4))
  stopifnot(vc$n34 <= min(vc$area3, vc$area4))
}

venn_counts_3way <- function(sets_named) {
  stopifnot(length(sets_named) == 3L)
  s1 <- sets_named[[1]]; s2 <- sets_named[[2]]; s3 <- sets_named[[3]]
  inter_n <- function(...) length(Reduce(intersect, list(...)))
  list(
    area1 = length(s1), area2 = length(s2), area3 = length(s3),
    n12 = inter_n(s1, s2), n13 = inter_n(s1, s3), n23 = inter_n(s2, s3),
    n123 = inter_n(s1, s2, s3)
  )
}

validate_venn_counts_3way <- function(vc) {
  stopifnot(all(unlist(vc) == as.integer(unlist(vc))))
  stopifnot(vc$n12 <= min(vc$area1, vc$area2))
  stopifnot(vc$n13 <= min(vc$area1, vc$area3))
  stopifnot(vc$n23 <= min(vc$area2, vc$area3))
  stopifnot(vc$n123 <= min(vc$n12, vc$n13, vc$n23))
}
```

```{r cross-study-overlap-5way}
##  broad external context: SCZD_PNN + habenulaPilot + refs 85/86/87(INT18).
set_names5 <- names(sets5)
membership5 <- build_membership_table(sets5)
combo_members_5way <- summarize_combos(membership5, set_names5)
combo_counts_5way <- combo_members_5way[, c("combo_label", "n_sets", set_names5, "n_genes"), with = FALSE]

f_5way_counts <- file.path(final_dir, "gene_list_overlap_5way_combo_counts.tsv.gz")
f_5way_members <- file.path(final_dir, "gene_list_overlap_5way_combo_members.tsv.gz")
fwrite(combo_counts_5way, f_5way_counts, sep = "\t", quote = FALSE)
fwrite(combo_members_5way, f_5way_members, sep = "\t", quote = FALSE)

small_intersections_5way <- combo_members_5way[
  (n_genes > 0L & n_genes < 20L) | combo_label == "SCZD_PNN",
  .(combo_label, n_sets, n_genes, gene_list)
]
setorder(small_intersections_5way, -n_genes, combo_label)
f_5way_small <- file.path(final_dir, "gene_list_overlap_5way_intersections_lt20.tsv.gz")
fwrite(small_intersections_5way, f_5way_small, sep = "\t", quote = FALSE)

vc5 <- venn_counts_5way(sets5)
validate_venn_counts_5way(vc5)

draw_venn5 <- function() {
  ##  label placement: 0=up, 90=right, 180=down, 270=left (per set in set_names5 order).
  ##  moved side labels (habenulaPilot and Wang2018_ref87_INT18) upward to avoid clipping.
  cat_pos5 <- c(0, 0, 215, 145, 0)
  cat_dist5 <- c(0.05, 0.12, 0.2, 0.2, 0.12)

  grid.newpage()
  v5 <- VennDiagram::draw.quintuple.venn(
    area1 = vc5$area1, area2 = vc5$area2, area3 = vc5$area3, area4 = vc5$area4, area5 = vc5$area5,
    n12 = vc5$n12, n13 = vc5$n13, n14 = vc5$n14, n15 = vc5$n15,
    n23 = vc5$n23, n24 = vc5$n24, n25 = vc5$n25, n34 = vc5$n34, n35 = vc5$n35, n45 = vc5$n45,
    n123 = vc5$n123, n124 = vc5$n124, n125 = vc5$n125, n134 = vc5$n134, n135 = vc5$n135,
    n145 = vc5$n145, n234 = vc5$n234, n235 = vc5$n235, n245 = vc5$n245, n345 = vc5$n345,
    n1234 = vc5$n1234, n1235 = vc5$n1235, n1245 = vc5$n1245, n1345 = vc5$n1345, n2345 = vc5$n2345,
    n12345 = vc5$n12345,
    category = set_names5,
    fill = c("#4f83b6", "#86b6d4", "#f4b183", "#a9d18e", "#b4a7d6"),
    alpha = rep(0.45, 5),
    cex = 0.85,
    cat.cex = 0.95,
    cat.pos = cat_pos5,
    cat.dist = cat_dist5,
    cat.col = c("#1f3f5b", "#1f3f5b", "#7f4a0c", "#3f6522", "#4a3a6c"),
    ind = FALSE
  )
  grid.draw(v5)
}

##  show in notebook output.
draw_venn5()

##  optional export for finalized layout tuning.
# f_venn5_pdf <- file.path(final_dir, "gene_list_overlap_venn_5way.pdf")
# pdf(f_venn5_pdf, width = 11, height = 8.5)
# draw_venn5()
# dev.off()

combo_counts_5way[n_genes > 0][order(-n_genes, combo_label)][1:min(.N, 15L)]
```

The broad 5-way overlap is useful for context, but it mixes adult DLPFC spatial colocalization, adult habenula bulk colocalization, adult multi-region brain colocalization candidates, fetal cortical colocalization, and an integrative PsychENCODE SCZ risk-prioritization resource. We therefore treat it as a broad external-context figure rather than a replication-style comparison.

For interpretation, the comparison tiers are: adult-focused external context (`SCZD_PNN`, `habenulaPilot`, `Zeng2022_ref85_SCZ`, `Wang2018_ref87_INT18`), developmental external context (`Wen2024_ref86`), and integrative SCZ-risk context (`Wang2018_ref87_INT18` with `INT17` as a stricter sensitivity subset).

```{r cross-study-overlap-3way-external-merged}
##  simplified summary: current study, habenula prior-study comparator, and merged external brain xQTL/GWAS support.
external_merged <- sort(unique(c(
  sets5$Zeng2022_ref85_SCZ,
  sets5$Wen2024_ref86,
  sets5$Wang2018_ref87_INT18
)))

sets3 <- list(
  SCZD_PNN = sets5$SCZD_PNN,
  habenulaPilot = sets5$habenulaPilot,
  External_Brain_xQTL_GWAS = external_merged
)

membership3 <- build_membership_table(sets3)
combo_members_3way <- summarize_combos(membership3, names(sets3))
combo_counts_3way <- combo_members_3way[, c("combo_label", "n_sets", names(sets3), "n_genes"), with = FALSE]

f_3way_counts <- file.path(final_dir, "gene_list_overlap_3way_external_merged_combo_counts.tsv.gz")
f_3way_members <- file.path(final_dir, "gene_list_overlap_3way_external_merged_combo_members.tsv.gz")
fwrite(combo_counts_3way, f_3way_counts, sep = "\t", quote = FALSE)
fwrite(combo_members_3way, f_3way_members, sep = "\t", quote = FALSE)

vc3 <- venn_counts_3way(sets3)
validate_venn_counts_3way(vc3)

draw_venn3 <- function() {
  grid.newpage()
  v3 <- VennDiagram::draw.triple.venn(
    area1 = vc3$area1, area2 = vc3$area2, area3 = vc3$area3,
    n12 = vc3$n12, n13 = vc3$n13, n23 = vc3$n23, n123 = vc3$n123,
    category = c("SCZD_PNN", "habenulaPilot", "External Brain xQTL/GWAS"),
    fill = c("#4f83b6", "#86b6d4", "#b4a7d6"),
    alpha = rep(0.45, 3),
    cex = 1.0,
    cat.cex = 0.95,
    cat.col = c("#1f3f5b", "#1f3f5b", "#4a3a6c"),
    ind = FALSE
  )
  grid.draw(v3)
}

##  show in notebook output.
draw_venn3()

##  optional export for finalized layout tuning.
# f_venn3_pdf <- file.path(final_dir, "gene_list_overlap_venn_3way_external_merged.pdf")
# pdf(f_venn3_pdf, width = 9, height = 7)
# draw_venn3()
# dev.off()

combo_counts_3way[n_genes > 0][order(-n_genes, combo_label)][]
```

This 3-way panel collapses the heterogeneous external references into a single umbrella comparison set for a cleaner high-level overlap view. The merged external circle combines adult brain eQTL/GWAS candidate genes, developmental xQTL/colocalization genes, and broad PsychENCODE SCZ-risk prioritization, so it should be interpreted as a generic external brain genetics comparison rather than a replication-only overlap.

```{r cross-study-overlap-sczd-unique}
##  SCZD_PNN-unique genes and their distribution across the 11 expression contexts.
sczd_unique <- membership5[
  SCZD_PNN & !habenulaPilot & !Zeng2022_ref85_SCZ & !Wen2024_ref86 & !Wang2018_ref87_INT18,
  .(gene_name, SCZD_PNN, habenulaPilot, Zeng2022_ref85_SCZ, Wen2024_ref86, Wang2018_ref87_INT18)
]
setorder(sczd_unique, gene_name)

f_sczd_unique <- file.path(final_dir, "gene_list_overlap_sczd_unique_genes.tsv.gz")
fwrite(sczd_unique, f_sczd_unique, sep = "\t", quote = FALSE)

sczd_pnn_coloc_strong <- unique(strong_dlpfc[, .(
  context = as.character(context_or_group),
  gene_name = std_symbol_key(gene_name),
  variant_id = as.character(variant_id),
  rsID = as.character(rsid)
)])
sczd_pnn_coloc_strong <- sczd_pnn_coloc_strong[!is.na(gene_name) & gene_name != ""]
sczd_pnn_coloc_strong[is.na(rsID) | rsID == "", rsID := ""]
setorder(sczd_pnn_coloc_strong, context, gene_name, variant_id, rsID)

f_sczd_pnn_coloc <- file.path(final_dir, "sczd-pnn-coloc-strong.csv")
fwrite(sczd_pnn_coloc_strong, f_sczd_pnn_coloc, sep = ",", quote = TRUE)

sczd_ctx <- unique(strong_dlpfc[, .(
  context = as.character(context_or_group),
  gene_name = std_symbol_key(gene_name)
)])
sczd_ctx <- sczd_ctx[!is.na(gene_name) & gene_name %chin% sczd_unique$gene_name]

sczd_ctx_summary <- sczd_ctx[, .(
  n_genes = uniqueN(gene_name),
  gene_list = paste(sort(unique(gene_name)), collapse = "; ")
), by = context]

sczd_ctx_summary <- merge(
  data.table(context = default_eqtl_contexts),
  sczd_ctx_summary,
  by = "context",
  all.x = TRUE
)
sczd_ctx_summary[is.na(n_genes), n_genes := 0L]
sczd_ctx_summary[is.na(gene_list), gene_list := ""]
sczd_ctx_summary[, context := factor(context, levels = default_eqtl_contexts)]
setorder(sczd_ctx_summary, context)
sczd_ctx_summary[, context := as.character(context)]

f_sczd_ctx <- file.path(final_dir, "gene_list_overlap_sczd_unique_contexts.tsv.gz")
fwrite(sczd_ctx_summary, f_sczd_ctx, sep = "\t", quote = FALSE)

sczd_ctx_summary
```

```{r cross-study-overlap-4way}
##  primary adult-focused overlap excluding the fetal reference set.
sets4 <- sets5[c("SCZD_PNN", "habenulaPilot", "Zeng2022_ref85_SCZ", "Wang2018_ref87_INT18")]
set_names4 <- names(sets4)
membership4 <- build_membership_table(sets4)
combo_members_4way <- summarize_combos(membership4, set_names4)
combo_counts_4way <- combo_members_4way[, c("combo_label", "n_sets", set_names4, "n_genes"), with = FALSE]

f_4way_counts <- file.path(final_dir, "gene_list_overlap_4way_combo_counts.tsv.gz")
f_4way_members <- file.path(final_dir, "gene_list_overlap_4way_combo_members.tsv.gz")
fwrite(combo_counts_4way, f_4way_counts, sep = "\t", quote = FALSE)
fwrite(combo_members_4way, f_4way_members, sep = "\t", quote = FALSE)

vc4 <- venn_counts_4way(sets4)
validate_venn_counts_4way(vc4)

draw_venn4 <- function() {
  grid.newpage()
  v4 <- VennDiagram::draw.quad.venn(
    area1 = vc4$area1, area2 = vc4$area2, area3 = vc4$area3, area4 = vc4$area4,
    n12 = vc4$n12, n13 = vc4$n13, n14 = vc4$n14, n23 = vc4$n23, n24 = vc4$n24, n34 = vc4$n34,
    n123 = vc4$n123, n124 = vc4$n124, n134 = vc4$n134, n234 = vc4$n234, n1234 = vc4$n1234,
    category = set_names4,
    fill = c("#4f83b6", "#f4b183", "#a9d18e", "#b4a7d6"),
    alpha = rep(0.45, 4),
    cex = 1.0,
    cat.cex = 1.0,
    cat.col = c("#1f3f5b", "#7f4a0c", "#3f6522", "#4a3a6c"),
    ind = FALSE
  )
  grid.draw(v4)
}

##  show in notebook output.
draw_venn4()

##  optional export for finalized layout tuning.
# f_venn4_pdf <- file.path(final_dir, "gene_list_overlap_venn_4way_adult_context.pdf")
# pdf(f_venn4_pdf, width = 10, height = 8)
# draw_venn4()
# dev.off()

combo_counts_4way[n_genes > 0][order(-n_genes, combo_label)][1:min(.N, 15L)]
```

```{r ref87-sensitivity}
##  sensitivity: compare the broad Ref87 list (INT18) with the high-confidence subset (INT17).
f_ref87_int17 <- file.path(hb_gene_dir, "Ref87_INT-17_SCZ_High_Confidence_Gene_List.csv")
assert_file(f_ref87_int17)

ref87_int17 <- fread(f_ref87_int17)
assert_cols(ref87_int17, c("sczgenenames"), "Ref87_INT-17_SCZ_High_Confidence_Gene_List.csv")

ref87_sets <- list(
  Wang2018_ref87_INT18 = to_gene_set(ref87_int18$sczgenenames),
  Wang2018_ref87_INT17 = to_gene_set(ref87_int17$sczgenenames)
)

ref87_sens_summary <- data.table(
  ref87_set = names(ref87_sets),
  n_genes = as.integer(lengths(ref87_sets)),
  n_overlap_sczd_pnn = vapply(ref87_sets, function(z) length(intersect(z, sets5$SCZD_PNN)), integer(1))
)

ref87_overlap_detail <- data.table(
  overlap_with_int18 = paste(intersect(ref87_sets$Wang2018_ref87_INT18, sets5$SCZD_PNN), collapse = "; "),
  overlap_with_int17 = paste(intersect(ref87_sets$Wang2018_ref87_INT17, sets5$SCZD_PNN), collapse = "; "),
  genes_lost_when_using_int17 = paste(setdiff(
    intersect(ref87_sets$Wang2018_ref87_INT18, sets5$SCZD_PNN),
    intersect(ref87_sets$Wang2018_ref87_INT17, sets5$SCZD_PNN)
  ), collapse = "; ")
)

stopifnot(all(ref87_sets$Wang2018_ref87_INT17 %chin% ref87_sets$Wang2018_ref87_INT18))

f_ref87_sens_sum <- file.path(final_dir, "gene_list_overlap_ref87_sensitivity_summary.tsv.gz")
f_ref87_sens_det <- file.path(final_dir, "gene_list_overlap_ref87_sensitivity_detail.tsv.gz")
fwrite(ref87_sens_summary, f_ref87_sens_sum, sep = "\t", quote = FALSE)
fwrite(ref87_overlap_detail, f_ref87_sens_det, sep = "\t", quote = FALSE)

ref87_sens_summary
ref87_overlap_detail
```

`Wang2018_ref87_INT17` is a stricter subset of `Wang2018_ref87_INT18`, but it reduces the already-small `SCZD_PNN` overlap. We therefore keep `INT18` as the broad Ref87 comparator and use `INT17` only as a sensitivity check.

```{r supplemental-prior-sensitivity}
##  load supplemental prior-sensitivity outputs if available and summarize strong-locus robustness.
if (file.exists(f_sens_sum) && file.exists(f_sens_gene)) {
  sens_sum <- fread(f_sens_sum)
  sens_gene <- fread(f_sens_gene)
  assert_cols(sens_gene, c("ds", "fid", "default_pass", "frac_pass", "min_pass_p12"), "prior sensitivity per-gene")

  strong_key <- unique(strong_dlpfc[, .(ds = context_or_group, fid = gene_id_raw)])
  sens_strong <- merge(
    strong_key,
    sens_gene,
    by = c("ds", "fid"),
    all.x = TRUE
  )
  sens_strong_summary <- sens_strong[, .(
    n_strong = .N,
    n_with_sensitivity = sum(!is.na(frac_pass)),
    n_default_pass = sum(default_pass == TRUE, na.rm = TRUE),
    mean_frac_pass = mean(frac_pass, na.rm = TRUE),
    n_any_pass_grid = sum(!is.na(min_pass_p12))
  ), by = ds][order(ds)]

  f_sens_strong <- file.path(final_dir, "strong_coloc_prior_sensitivity_summary.tsv.gz")
  fwrite(sens_strong_summary, f_sens_strong, sep = "\t")
  sens_strong_summary
} else {
  message("prior-sensitivity files not found; skipping supplemental panel.")
}
```

```{r supplemental-gate-failures}
##  load and summarize gate outcome reasons by category and context in a compact panel.
if (file.exists(f_sens_fail) && file.exists(f_sens_fail_table)) {
  sens_fail <- fread(f_sens_fail)
  sens_fail_table <- fread(f_sens_fail_table)
  fail_compact <- sens_fail[, .(n_loci = sum(n_loci)), by = .(cat_raw, gate_reason)][order(cat_raw, gate_reason)]

  f_fail_compact <- file.path(final_dir, "prior_sensitivity_gate_reason_counts.tsv.gz")
  f_fail_by_context <- file.path(final_dir, "prior_sensitivity_gate_reason_by_context.tsv.gz")
  fwrite(fail_compact, f_fail_compact, sep = "\t")
  fwrite(sens_fail_table, f_fail_by_context, sep = "\t")
  fail_compact
  sens_fail_table

  fail_plot <- copy(sens_fail_table)
  fail_plot[, ds := factor(ds, levels = sort(unique(ds)))]
  fail_plot[, cat_raw := factor(cat_raw, levels = c("strong_coloc", "likely_coloc", "distinct_causal", "follow_up"))]

  ggplot(fail_plot, aes(x = ds, y = n_loci, fill = gate_reason)) +
    geom_col(width = 0.75) +
    coord_flip() +
    facet_wrap(~ cat_raw, ncol = 1, scales = "free_y") +
    theme_bw() +
    labs(
      x = "context",
      y = "number of loci",
      title = "prior-sensitivity gate outcomes by context",
      subtitle = "mid-level summary by cat_raw and gate_reason"
    ) +
    scale_fill_manual(values = c(
      pass = "#6aa84f",
      fail_default_rule = "#cc0000",
      fail_cat_instability = "#e69138",
      fail_no_plausible_points = "#8e7cc3",
      fail_sensitivity_error = "#999999",
      not_evaluated = "#c9c9c9"
    ))
} else {
  message("prior-sensitivity fail summary files not found; skipping gate-failure panel.")
}
```

```{r mini-coloc-kansl1-as1-spd04}
##  compact regional-association panel for the SpD04-WM KANSL1-AS1 coloc locus.
f_mini_gwas <- file.path(refdir, "GWAS_SCZ_tqtl-matched.tab.gz")
f_mini_eqtl <- file.path(datadir, "tqtl_out", "spd04.gene.cis_qtl_pairs.chr17.parquet")
f_mini_coloc <- file.path(datadir, "coloc", "coloc_spd04_gene.qs2")
f_mini_gated <- file.path(datadir, "coloc", "coloc_abf_results_gated.tsv.gz")
f_mini_exprbed <- file.path(datadir, "tqtl_in", "spd04.gene.expr.bed.gz")
assert_file(f_mini_gwas)
assert_file(f_mini_eqtl)
assert_file(f_mini_coloc)
assert_file(f_mini_gated)
assert_file(f_mini_exprbed)

mini_context <- "spd04"
mini_context_label <- "SpD04-WM"
mini_gene_id <- "ENSG00000214401"
mini_gene_symbol <- "KANSL1-AS1"
mini_lead_variant <- "chr17:46188473:C:T"
mini_lead_rsid <- "rs2532240"
mini_window_left_bp <- 300000L
mini_window_right_bp <- 125000L
mini_gw_line <- -log10(5e-8)

mini_gated <- fread(f_mini_gated)
mini_gated_row <- mini_gated[
  ds == mini_context & fid == mini_gene_id & lead_snp == mini_lead_variant
]
if (nrow(mini_gated_row) != 1L) {
  stop("Expected exactly one gated coloc row for the SpD04-WM KANSL1-AS1 locus.")
}

mini_coloc_obj <- qs2::qs_read(f_mini_coloc)[[mini_gene_id]]
if (is.null(mini_coloc_obj) || is.null(mini_coloc_obj$results) || is.null(mini_coloc_obj$summary)) {
  stop("Missing coloc results for the SpD04-WM KANSL1-AS1 locus.")
}
mini_coloc_res <- as.data.table(mini_coloc_obj$results)
assert_cols(mini_coloc_res, c("snp", "SNP.PP.H4"), "mini coloc SNP-level results")
mini_coloc_res[, c("chr", "pos", "ref", "alt") := tstrsplit(snp, ":", fixed = TRUE)]
mini_coloc_res[, pos := as.integer(pos)]

mini_gwas_dt <- fread(f_mini_gwas, select = c("variant_id", "rsid", "p", "beta", "beta_se"))
mini_gwas_dt[, c("chr", "pos", "ref", "alt") := tstrsplit(variant_id, ":", fixed = TRUE)]
mini_gwas_dt[, pos := as.integer(pos)]

mini_eqtl_dt <- as.data.table(arrow::read_parquet(
  f_mini_eqtl,
  col_select = c("phenotype_id", "variant_id", "start_distance", "pval_nominal", "slope", "slope_se")
))
mini_eqtl_dt <- mini_eqtl_dt[phenotype_id == mini_gene_id]
mini_eqtl_dt[, c("chr", "pos", "ref", "alt") := tstrsplit(variant_id, ":", fixed = TRUE)]
mini_eqtl_dt[, pos := as.integer(pos)]

mini_exprbed <- fread(f_mini_exprbed, select = 1:4)
setnames(mini_exprbed, c("#Chr", "start", "end", "ID"), c("chr", "start", "end", "gene_id"))
mini_gene_row <- mini_exprbed[gene_id == mini_gene_id]
if (nrow(mini_gene_row) != 1L) {
  stop("Expected exactly one phenotype BED row for the SpD04-WM KANSL1-AS1 locus.")
}
mini_tss_pos <- mini_gene_row$start[1]

mini_lead_pos <- mini_gwas_dt[variant_id == mini_lead_variant, pos][1]
if (is.na(mini_lead_pos)) stop("Lead variant not found in harmonized GWAS table.")

mini_gwas_dt <- mini_gwas_dt[
  chr == "chr17" &
    pos >= mini_lead_pos - mini_window_left_bp &
    pos <= mini_lead_pos + mini_window_right_bp
]
mini_eqtl_dt <- mini_eqtl_dt[
  chr == "chr17" &
    pos >= mini_lead_pos - mini_window_left_bp &
    pos <= mini_lead_pos + mini_window_right_bp
]

mini_eqtl_dt <- merge(
  mini_eqtl_dt,
  mini_gwas_dt[, .(variant_id, rsid)],
  by = "variant_id",
  all.x = TRUE,
  suffixes = c("", "_gwas")
)
mini_eqtl_dt <- merge(
  mini_eqtl_dt,
  mini_coloc_res[, .(variant_id = snp, SNP.PP.H4)],
  by = "variant_id",
  all.x = TRUE
)

mini_eqtl_dt[, rsid := fifelse(!is.na(rsid) & rsid != "", rsid, variant_id)]
mini_eqtl_dt[, track := "eQTL"]
mini_eqtl_dt[, neglog10p := -log10(pval_nominal)]
mini_eqtl_dt[, is_lead := variant_id == mini_lead_variant]
mini_eqtl_dt[, pos_mb := pos / 1e6]

mini_gwas_dt[, track := "GWAS"]
mini_gwas_dt[, neglog10p := -log10(p)]
mini_gwas_dt[, is_lead := variant_id == mini_lead_variant]
mini_gwas_dt[, pos_mb := pos / 1e6]

mini_coloc_meta <- data.table(
  context = mini_context,
  context_label = mini_context_label,
  gene_id = mini_gene_id,
  gene_symbol = mini_gene_symbol,
  lead_variant = mini_lead_variant,
  lead_rsid = mini_lead_rsid,
  lead_pos = mini_lead_pos,
  lead_pos_mb = mini_lead_pos / 1e6,
  tss_pos = mini_tss_pos,
  tss_pos_mb = mini_tss_pos / 1e6,
  rsid_validated = mini_gwas_dt[variant_id == mini_lead_variant, rsid][1] == mini_lead_rsid,
  cat_gated = mini_gated_row$cat_gated[1],
  gate_reason = mini_gated_row$gate_reason[1],
  PP4 = mini_gated_row$PP4[1],
  PP34 = mini_gated_row$PP34[1],
  PP4_over_PP34 = mini_gated_row$PP4_over_PP34[1],
  lead_to_tss_bp = mini_lead_pos - mini_tss_pos,
  lead_snp_pp_h4 = mini_eqtl_dt[is_lead == TRUE, SNP.PP.H4][1],
  top3_snp_pp_h4_sum = mini_coloc_res[order(-SNP.PP.H4)][1:min(.N, 3L), sum(SNP.PP.H4)]
)

if (!isTRUE(mini_coloc_meta$rsid_validated[1])) {
  stop("Lead rsID mismatch between the harmonized GWAS table and expected mini coloc locus.")
}

mini_plot_dt <- rbindlist(list(
  mini_eqtl_dt[, .(track, pos_mb, neglog10p, is_lead, label = fifelse(is_lead, mini_lead_rsid, NA_character_))],
  mini_gwas_dt[, .(track, pos_mb, neglog10p, is_lead, label = fifelse(is_lead, mini_lead_rsid, NA_character_))]
), use.names = TRUE)
mini_plot_dt[, track := factor(track, levels = c("eQTL", "GWAS"))]

mini_label_dt <- mini_plot_dt[is_lead == TRUE]
mini_label_dt[, y_label := neglog10p + 0.06 * max(mini_plot_dt[track == .BY$track, neglog10p], na.rm = TRUE), by = track]
mini_tss_dt <- data.table(
  track = factor("eQTL", levels = levels(mini_plot_dt$track)),
  xintercept = mini_tss_pos / 1e6
)
mini_hline_dt <- data.table(track = factor("GWAS", levels = levels(mini_plot_dt$track)), yint = mini_gw_line)
mini_x_limits <- c(
  mini_lead_pos - mini_window_left_bp,
  mini_lead_pos + mini_window_right_bp
) / 1e6

p_kansl1_as1_spd04_wm_mini <- ggplot(mini_plot_dt, aes(x = pos_mb, y = neglog10p)) +
  geom_vline(
    data = mini_tss_dt,
    aes(xintercept = xintercept),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.35,
    color = "#666666"
  ) +
  geom_hline(
    data = mini_hline_dt,
    aes(yintercept = yint),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 0.3,
    color = "#6b4c6f"
  ) +
  geom_point(
    data = mini_plot_dt[is_lead == FALSE],
    aes(color = track),
    size = 1.5,
    alpha = 0.9
  ) +
  geom_point(
    data = mini_plot_dt[is_lead == TRUE],
    color = "#d7301f",
    size = 2.2
  ) +
  geom_text(
    data = mini_label_dt,
    aes(y = y_label, label = label),
    color = "#d7301f",
    size = 2.8,
    fontface = "bold",
    vjust = 0
  ) +
  facet_grid(track ~ ., scales = "free_y", switch = "y") +
  coord_cartesian(xlim = mini_x_limits, clip = "off") +
  scale_color_manual(values = c(eQTL = "#eb8c23", GWAS = "#3b84c6")) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.16))) +
  labs(
    x = "Chromosomal position (Mb)",
    y = expression(-log[10](P)),
    title = "Colocalization signal for KANSL1-AS1",
    subtitle = sprintf(
      "%s; lead SNP %s; coloc PP4 = %.2f",
      mini_context_label,
      mini_lead_rsid,
      mini_coloc_meta$PP4[1]
    )
  ) +
  theme_bw(base_size = 8) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 10),
    plot.subtitle = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    strip.text.y.left = element_text(angle = 0, face = "bold", size = 8),
    strip.placement = "outside",
    panel.spacing.y = unit(0.15, "lines"),
    plot.margin = margin(t = 5.5, r = 18, b = 5.5, l = 5.5)
  )

f_mini_pdf <- file.path(final_dir, "kansl1_as1_spd04_wm_mini_coloc_panel.pdf")
f_mini_png <- file.path(final_dir, "kansl1_as1_spd04_wm_mini_coloc_panel.png")
ggsave(f_mini_pdf, p_kansl1_as1_spd04_wm_mini, width = 3.2, height = 2.6, units = "in", dpi = 300)
ggsave(f_mini_png, p_kansl1_as1_spd04_wm_mini, width = 3.2, height = 2.6, units = "in", dpi = 300)

mini_coloc_meta
p_kansl1_as1_spd04_wm_mini
```

The stacked `GWAS` and `eQTL` tracks are a compact regional-association view of concordant signal at the `KANSL1-AS1` locus in `SpD04-WM`; the colocalization inference itself is still the posterior support (`PP4 = 0.88`), and `rs2532240` should be interpreted as the top shared-signal candidate rather than a uniquely proven causal SNP.
